---
title: "Clustering Exercise: Customer Segmentation"
author: "MKT 566 · Decision Making Using Marketing Analytics · Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
---

<!-- Student-facing: all code run, figures and tables displayed, questions
     kept, answers removed. Generated from the same body as
     w6-clustering-exe-solution.Rmd. Reads customer_clustering_data.csv
     from its own folder. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
library(data.table)
library(ggplot2)
options(width = 110)
```

**Name and USC ID:** ______

The coding is done: every chart and table below is produced by the code
shown above it. Your job is **the reading**. For each task, answer the
questions in plain English, with the number that supports each claim, in the
**My answers:** block under the questions.

Using your AI assistant to talk through a table is fine, and a good idea. The
sentences you write must be your own reading.

## How to work on it

1. Download the [R Markdown source of this page](https://github.com/dadepro/mkt566/raw/main/w6/case/w6-clustering-exe.Rmd)
   (right-click, Save Link As...) and the
   [data](https://github.com/dadepro/mkt566/raw/main/w6/case/clustering-exercise-data.zip).
   Put `w6-clustering-exe.Rmd` and the unzipped `customer_clustering_data.csv`
   in the same folder, and open that folder in VS Code.
2. Fill in every **My answers:** block, and your name above.
3. Knit to HTML with the play button and check that every figure and table
   is visible in the knitted file.

## The data

One row per **customer** of a retailer that sells both online and in its
stores. Columns:

| Column | What it is |
|---|---|
| `CustomerID` | the customer |
| `Age` | age in years |
| `Income` | annual income, in dollars |
| `Online_Spend` | dollars the customer spent online last year |
| `InStore_Spend` | dollars the customer spent in the stores last year |

The retailer wants to know **what kinds of customers it has**, and what to
do with each kind. There is no outcome to predict: this is k-means and PCA,
as in Tuesday's slides.

## Task 1: look at the data

```{r task1}
# load the data (the file must be in the same folder as this Rmd)
df <- fread("customer_clustering_data.csv")

# the first rows, and a summary of the four numeric columns
head(df)
summary(df[, .(Age, Income, Online_Spend, InStore_Spend)])
```

- How many customers are there, and what does one row describe?
- Look at the ranges of the four columns. Which column has by far the
  largest numbers? Why does that matter for k-means? (Tuesday's "Watch the
  units" slide.)

**My answers:**

## Task 2: a first picture

```{r task2}
ggplot(df, aes(x = Online_Spend, y = InStore_Spend)) +
  geom_point(size = 2, alpha = 0.7, color = "grey35") +
  labs(title = "Online vs. In-Store Spend, Last Year",
       x = "Online spend ($)", y = "In-store spend ($)") +
  theme_minimal(base_size = 13)
```

- How many groups do you see? Describe each one in words.

**My answers:**

## Task 3: standardize, then choose k

```{r task3}
# the four columns we cluster on
feat <- df[, .(Online_Spend, InStore_Spend, Age, Income)]

# standardize: each column becomes (value - its mean) / its standard deviation
X <- scale(feat)
round(head(X), 2)

# elbow curve: run k-means for k = 2 to 10 and record the within-cluster spread
set.seed(123)
wss <- sapply(2:10, function(k) kmeans(X, centers = k, nstart = 20)$tot.withinss)
elbow <- data.table(k = 2:10, wss = round(wss, 1))
elbow

ggplot(elbow, aes(x = k, y = wss)) +
  geom_line(color = "#990000", linewidth = 1.1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 2:10) +
  labs(title = "Total Within-Cluster Sum of Squares vs. k",
       x = "Number of clusters k", y = "Total within-cluster SS") +
  theme_minimal(base_size = 13)
```

- After `scale()`, what does a value of 0 in `X` mean? And a value of 1?
- Where is the elbow? Propose a k, with the numbers from the `elbow` table.

**My answers:**

## Task 4: k-means with k = 3, and the profile table

```{r task4}
# k-means on the standardized columns; nstart = 20 tries 20 random starts
set.seed(123)
km <- kmeans(X, centers = 3, nstart = 20)
df[, cluster := factor(km$cluster)]

# profile each cluster in the ORIGINAL units (dollars, years)
profile <- df[, .(customers     = .N,
                  online_spend  = round(mean(Online_Spend)),
                  instore_spend = round(mean(InStore_Spend)),
                  total_spend   = round(mean(Online_Spend + InStore_Spend)),
                  age           = round(mean(Age), 1),
                  income        = round(mean(Income)),
                  total_all     = sum(Online_Spend + InStore_Spend)),
              keyby = cluster]
# each cluster's share of all the money customers spent
profile[, share_of_spend_pct := round(100 * total_all / sum(total_all), 1)]
profile[, total_all := NULL]
profile

ggplot(df, aes(x = Online_Spend, y = InStore_Spend, color = cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = c("1" = "#990000", "2" = "#0072B2", "3" = "#E69F00")) +
  labs(title = "The Same Customers, Colored by K-Means (k = 3)",
       x = "Online spend ($)", y = "In-store spend ($)", color = "Cluster") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

- Describe each cluster with the numbers from the profile table, and give
  each one a name a marketer would use.
- Why is the profile table computed on the original columns (`df`) and not
  on the standardized ones (`X`)?

**My answers:**

## Task 5: what if we had not standardized?

```{r task5}
# the same k-means, on the raw columns (no scale())
set.seed(123)
km_raw <- kmeans(feat, centers = 3, nstart = 20)
df[, cluster_raw := factor(km_raw$cluster)]

df[, .(customers     = .N,
       online_spend  = round(mean(Online_Spend)),
       instore_spend = round(mean(InStore_Spend)),
       age           = round(mean(Age), 1),
       income        = round(mean(Income))),
   keyby = cluster_raw]

# where did each customer go? Rows: the raw clusters from this task.
# Columns: the standardized clusters from Task 4. Each cell counts the
# customers who are in both.
where <- as.data.frame.matrix(table(df$cluster_raw, df$cluster))
colnames(where) <- paste("Task 4 cluster", colnames(where))
rownames(where) <- paste("Raw cluster", rownames(where))
knitr::kable(where, align = "c",
             caption = "Number of customers in each raw cluster (rows) and each Task 4 cluster (columns)")
```

- What separates the three raw clusters? Look at the income column.
- Use the last table (where each customer went): which Task 4 cluster gets
  split or mixed up? Which
  clustering would you hand to the marketing team?

**My answers:**

## Task 6: the PCA map

```{r task6}
# PCA on the four columns, standardized inside prcomp()
pca <- prcomp(feat, center = TRUE, scale. = TRUE)
summary(pca)

# loadings: how much each column counts in PC1 and PC2
round(pca$rotation[, 1:2], 2)

# the map: every customer at its (PC1, PC2), colored by cluster, plus the arrows
scores   <- data.table(pca$x[, 1:2], cluster = df$cluster)
loadings <- data.table(pca$rotation[, 1:2], variable = rownames(pca$rotation))
arrow_scale <- 2.5

ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  geom_point(data = scores, aes(x = PC1, y = PC2, color = cluster),
             size = 2, alpha = 0.7) +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1 * arrow_scale, yend = PC2 * arrow_scale),
               arrow = arrow(length = unit(0.25, "cm")), linewidth = 0.8) +
  geom_label(data = loadings,
             aes(x = PC1 * arrow_scale * 1.3, y = PC2 * arrow_scale * 1.3,
                 label = variable), size = 4, fill = "white", alpha = 0.85) +
  scale_color_manual(values = c("1" = "#990000", "2" = "#0072B2", "3" = "#E69F00")) +
  labs(title = "PCA Map of the 198 Customers",
       x = "PC1", y = "PC2", color = "Cluster") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

- How much of the variation do PC1 and PC2 keep together? (The
  "Cumulative Proportion" row of the summary.)
- Read the loadings: what does it mean to move **right** on PC1? To move
  **down** on PC2? (PCA picks the direction of each axis arbitrarily; only
  the arrows tell you what the axes mean.)
- Where does each cluster sit on the map? Do they separate?

**My answers:**

## Task 7: the recommendation

- Which segment is the most valuable: per customer, and in total? Use the
  profile table from Task 4.
- Propose **one marketing action per segment**: which channel, which offer,
  which message.
- State **one limitation** of this analysis for those decisions.

**My answers:**
