---
title: 'Exploratory data analysis: RateBeer dataset'
output:
  html_document:
    df_print: paged
---

```{r, message=FALSE, warning=FALSE}
#load libraries
library(data.table)
library(scales)
library(ggplot2)
library(stargazer)
library(ggthemes)
library(here)
library(forcats)
# set the working directory to be the location of your data file
setwd("/Users/dproserp/Library/CloudStorage/Dropbox/teaching/mkt566-2025/w3/beer-case")
```

In this exercise, we will work with a dataset of beer reviews from RateBeer (you can download it [here](insert_link)), and perform some exploratory data analysis.

Let's start by exploring the data and familiarizing with its structure:

```{r, message=FALSE, warning=FALSE}
beer = fread("w3-ratebeer-sampled.csv.gz")
head(beer)
```

1.	How many variables are in the data?
```{r}
ncol(beer)
```
 
2.	How many rows?
```{r}
nrow(beer)
```

Now, let’s focus on some specific variables:

3.	Explore the values of beer_style, do you see anything strange or unusual?

**SOLUTION**: Yes, there are non-beer drinks in the dataset, such as "Cider" and "Mead". We will need to filter these out later.
```{r}
unique(beer$beer_style)
```

4.	Explore the beer_ABV variable, do you see anything unusual?

**SOLUTION**: Yes, there are some values that are not numeric, such as "_". Also, there are beers with very high ABV values.

5.	What does the "_" value mean in the beer_ABV variable?

**SOLUTION**: The "_" character indicates that the ABV value is missing but it does not seem to be zero (since non-alcholic beer have ABV). We will need to handle these missing values appropriately.

6.	Explore the rating (review_XXX) variables. Are you able to compute the mean? If not, what is the issue?

**SOLUTON**: The rating variables (review_overall, review_taste, etc.) are currently stored as character strings, which prevents us from computing the mean directly. We need to convert them to numeric format first.

7.	Explore the variable review_time. What do you notice? How can you convert it to a date format?

**SOLUTION**: The review_time variable is in Unix timestamp format, which represents the number of seconds since January 1, 1970. We can convert it to a date format using the `as.POSIXct` function in R.


Now that we are familiar with the data, let’s try to visualize it and answer some questions. Before starting, remove non-beers drinks from the dataset, and convert all variables with prefix “review_” and beer_ABV to numerical variables.

```{r, message=FALSE, warning=FALSE}
# fix a few things

#conver timestamp to datatime
beer[, datetime:= as.POSIXct(review_time, origin = "1970-01-01", tz = "UTC")]
#remove non-beers drinks
beer = beer[!grepl("Sak|Cider|Mead|Kombucha|Wine|Liquor", beer_style)]
#convert all review_ variables to numeric
for (col in c("review_overall", "review_aroma", "review_appearance", "review_palate", "review_taste")) {
  beer[, (col) := as.numeric(sub("/.*", "", get(col)))]
}
# conver abv to numeric
beer[, beer_ABV := as.numeric(beer_ABV)]
```

8.	Plot the (ordered) distribution of reviews by beer_style. Which one is the most and least popular beer?

```{r, message=FALSE, warning=FALSE}
beer$beer_style <- fct_infreq(beer$beer_style)
ggplot(data = beer) +
  geom_bar(mapping = aes(x = beer_style), fill = "blue", color = "black") +
  labs(title = "Distribution of Beer Styles", x = "Beer Style", y = "Reviews") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size= 6),
        plot.margin  = margin(t = 10, r = 20, b = 40, l = 20)) +
  scale_y_continuous(labels = comma_format()) 
```

9.	Plot the distribution of beer_ABV for all beers. Does it look like you would expect? Are there any outliers? Which one is the beer with the highest ABV?

```{r, message=FALSE, warning=FALSE}
ggplot(data = beer) +
  geom_histogram(mapping = aes(x = beer_ABV), binwidth = 0.5, fill = "blue", color = "black") +
  labs(title = "Distribution of ABV", x = "ABV", y = "Count") +
  scale_x_continuous(breaks = seq(0, max(beer$beer_ABV, na.rm = TRUE), 2)) +
  theme_minimal()
```
```{r, message=FALSE, warning=FALSE}
# Example: Find outliers in the ABV variable
Q1 <- quantile(beer$beer_ABV, 0.25, na.rm = TRUE)
Q3 <- quantile(beer$beer_ABV, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1
# Outliers are values below Q1 - 1.5*IQR or above Q3 + 1.5*IQR
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

outliers = unique(beer[beer_ABV < lower_bound | beer_ABV > upper_bound, .(beer_name, beer_ABV)])
outliers
```

```{r, message=FALSE, warning=FALSE}
# beer with max ABV
max_abv_beer <- beer[which.max(beer$beer_ABV), .(beer_name, beer_style, beer_ABV)]
max_abv_beer
```




10.	Plot the distribution of average beer_ABV by beer_style. Which beer_style has the most and least ABV content?
```{r, message=FALSE, warning=FALSE}
avg.abv = beer[, .(beer_ABV = mean(as.numeric(beer_ABV), na.rm = TRUE)), by = beer_style]
ggplot(data = avg.abv) +
  geom_bar(mapping = aes(x = reorder(beer_style, -beer_ABV), y = beer_ABV), stat = "identity") +
  labs(title = "Average ABV by Beer Style", x = "Beer Style", y = "Average ABV") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  scale_y_continuous(labels = scales::percent_format(scale = 1))
```

Next, let’s look at ratings:

11.	Plot the average review_overall by beer_style. Which one is the highest- and lowest-rated style?

```{r, message=FALSE, warning=FALSE}
avg.review = beer[, .(review_overall = mean(review_overall, na.rm = TRUE)), by = beer_style]
ggplot(data = avg.review, aes(x = reorder(beer_style, -review_overall), y = review_overall, group = 1)) +
  geom_line() + geom_point() +
  labs(title = "Average Review Overall by Beer Style", x = "Beer Style", y = "Average Review Overall") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  scale_y_continuous(labels = scales::number_format(scale = 1))
```

12.	Plot the average review_taste by beer_style. Which one is the highest- and lowest-rated style?
```{r, message=FALSE, warning=FALSE}
avg.taste = beer[, .(review_taste = mean(review_taste, na.rm = TRUE)), by = beer_style]
ggplot(data = avg.taste, aes(x = reorder(beer_style, -review_taste), y = review_taste, group = 1)) +
  geom_line() + geom_point() +
  labs(title = "Average Review Taste by Beer Style", x = "Beer Style", y = "Average Review Taste") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  scale_y_continuous(labels = scales::number_format(scale = 1))
```

13.	Plot the distribution of average review_overall by beer_brewerId. Do you see anything unusual? Which brewery is the most and least liked, with at least 10 reviews?

**SOLUTION**: There are spikes concentrated in roung numbers (e.g., 2, 3, 4, 4.5, 5), which is due to ratings being round numbers and averages more like to have specific values that  are round number of multiples of 0.05 in most cases

```{r, message=FALSE, warning=FALSE}
avg.brewer = beer[, .(review_overall = mean(review_overall, na.rm = TRUE), num_reviews = .N), by = beer_brewerId]
ggplot(data = avg.brewer, aes(x = review_overall)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "black") +
  labs(title = "Distribution of Average Review Overall by Brewer", x = "Average Review Overall", y = "Count") +
  scale_x_continuous() +
  theme_minimal()
#Which are the best and worst brewer?
best.brewer = avg.brewer[review_overall == max(review_overall)]
worst.brewer = avg.brewer[review_overall == min(review_overall)]
best.brewer  
worst.brewer
# with at least 10 reviews
best.brewer = avg.brewer[num_reviews >= 10][review_overall == max(review_overall)]
worst.brewer = avg.brewer[num_reviews >= 10][review_overall == min(review_overall)]
best.brewer
worst.brewer
```

Finally, let’s look at the relationship between beer styles and reviews:

14.	Are breweries with a few or many beer styles those with higher reviews_overall?
```{r, message=FALSE, warning=FALSE}
avg.brewer.style = beer[, .(num_styles = uniqueN(beer_style), review_overall = mean(review_overall, na.rm = TRUE)), by = beer_brewerId]
ggplot(data = avg.brewer.style, aes(x = num_styles, y = review_overall)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Number of Beer Styles vs. Average Review Overall", x = "Number of Beer Styles", y = "Average Review Overall") +
  scale_x_continuous(breaks = seq(0, max(avg.brewer.style$num_styles, na.rm = TRUE), 5)) +
  theme_minimal()
```

Alternatively:
```{r, message=FALSE, warning=FALSE}
# compute 
temp = beer[, list(num_beer_style =length(unique(beer_style)),
                  avg_rating = mean(review_overall, na.rm=T)), by = beer_brewerId]
#compute average review_overall by num_beer_style
avg.brewer.style = temp[, .(avg_rating = mean(avg_rating, na.rm = TRUE)), 
                         by = num_beer_style][order(-num_beer_style)]

ggplot(data = avg.brewer.style, aes(x = num_beer_style, y = avg_rating)) +
  geom_line() + geom_point() +
  labs(title = "Average Review Overall by Number of Beer Styles", x = "Number of Beer Styles", y = "Average Review Overall") +
  scale_x_continuous(breaks = seq(0, max(avg.brewer.style$num_beer_style), 5)) +
  scale_y_continuous(labels = scales::number_format(scale = 1)) +
  theme_minimal()
```

15. Suppose you were tasked with helping a brewery decide which and how many beers to produce. How can you use this data to answer these questions? You can utilize some of the analyses you performed above and also add new ones (for example, we did not examine time trends). Prepare a brief presentation that discusses the analyses supporting your decisions.
