---
title: "Clustering Exercise: Customer Segmentation"
author: "Davide Proserpio"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
set.seed(123)
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 7, fig.height = 5)

# Load packages
library(ggplot2)
library(data.table)
library(readxl)
library(ggrepel)
library(skimr)
```

## Load the Data

```{r load}
f <- "customer_clustering_data.csv"
df <- fread(f)
head(df)
print(skim(df))
```

## Quick EDA (Scatter & Distributions)

Let's plot the scatter plot of Online vs In-Store Spend, income, and age distribution
```{r eda}
p1 <- ggplot(df, aes(Online_Spend, InStore_Spend)) +
  geom_point(alpha = 0.6) +
  labs(title = "Online vs In-Store Spend", x = "Online Spend", y = "In-Store Spend") +
  theme_minimal()

p2 <- ggplot(df, aes(Income)) +
  geom_histogram(bins = 30, fill = "grey70", color = "white") +
  labs(title = "Income Distribution", x = "Income", y = "Count") +
  theme_minimal()

p3 <- ggplot(df, aes(Age)) +
  geom_histogram(bins = 30, fill = "grey70", color = "white") +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()

p1
p2
p3
```

- Do you see natural groupings by spend/income/age (e.g., online-heavy, in-store-heavy, omni-channel)?

**SOLUTION:** Yes: clusters around high in-store / low online; high online / low in-store; and high both (omni-channel).

## Prepare Features & Scale

```{r scale}
features <- df[, .(Online_Spend, InStore_Spend, Age, Income)]
X <- scale(features)  # mean=0, sd=1
summary(X)
```

- Why do we scale before k-means?

**SOLUTION:** K-means uses Euclidean distance; unscaled high-variance features (Income/Spend) would dominate distance and clustering.

## Choose k: Elbow Method (WSS)

```{r elbow}
wss <- sapply(2:10, function(k){
  kmeans(X, centers = k, nstart = 25)$tot.withinss
})
## plot the within-cluster sum of squares
elbow_plot <- data.frame(k = 2:10, wss = wss)
ggplot(elbow_plot, aes(x = k, y = wss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2:10) +
  labs(title = "Elbow Method: Total Within-Cluster Sum of Squares vs. Number of Clusters",
       x = "Number of Clusters (k)",
       y = "Total Within-Cluster Sum of Squares") +
  theme_minimal()
```

- Inspect the plot. Where is the “elbow”? Propose a k.

**SOLUTION** Elbow typically around k ≈ 3 for this dataset.



## Fit K-means (k = 3)

```{r kmeans-fit}
k <- 3
km <- kmeans(X, centers = k, nstart = 50)

df$cluster <- factor(km$cluster)

# Cluster visualization in original spend space
ggplot(df, aes(Online_Spend, InStore_Spend, color = cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = paste("K-means Clusters (k =", k, ")"),
       x = "Online Spend", y = "In-Store Spend") +
  theme_minimal() +
  theme(legend.position = "top")
```

- What business personas do these clusters represent?

**SOLUTION** (1) Online-heavy low in-store; (2) In-store-heavy low online; (3) High both (omni-channel, likely highest value).

## Cluster Profiling (Means by Cluster)

```{r profiling}
profile = df[, .(Online_Spend = mean(Online_Spend),
                 InStore_Spend = mean(InStore_Spend),
                 Age = mean(Age),
                 Income = mean(Income)),
             by = cluster]
profile
```

- Which cluster looks most valuable? What actions would you take per segment?

**SOLUTION** Omni-channel (high online & in-store) likely most valuable (cross-sell, loyalty). Online-heavy: digital promotions; In-store-heavy: local offers/merchandising.


## PCA Projection

```{r pca}
pca_result <- prcomp(X, center = FALSE, scale. = FALSE)
# check variance explained
summary(pca_result)
```

- How many PCs explain >80% of variance?

### Extract PCA Loadings

```{r pca-loadings}
# Keep first two principal components (scores) and add cluster labels
pca_data <- data.frame(pca_result$x[, 1:2], cluster = df$cluster)

# Extract loadings which tell you how each original variable contributes to each PC.
pca_loadings <- as.data.frame(pca_result$rotation[, 1:2]) # 2 columns: PC1, PC2
pca_loadings$variable <- rownames(pca_loadings)           # keep var names for plotting
pca_loadings[,1:2]
```

- What do the loadings tell you about how each original variable contributes to each PC?

**SOLUTION** PC1: positive loadings on In-Store Spend, Age, Income; PC2: negative loading on Online_Spend, positive on age, negative on income.

### Plot PCA loadings

```{r pca-loadings-plot}
#plot loadings
# Scale loadings for better visibility in plots
arrow_scale <- 5   # tweak if arrows are too short/long

# Plot loadings alone (annotated with variables names)
p_loadings <- ggplot(pca_loadings, aes(x = PC1 * arrow_scale, y = PC2 * arrow_scale, label = variable)) +
  geom_segment(aes(x = 0, y = 0, xend = PC1 * arrow_scale, yend = PC2 * arrow_scale),
               arrow = arrow(length = unit(0.3, "cm")), color = "blue") +
  geom_text(size = 5) +
  labs(
    title = "PCA Loadings: Store attributes",
    x = "\nPrincipal Component 1",
    y = "Principal Component 2\n"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  theme_minimal() +
  coord_cartesian(xlim = c(-2, 8), ylim = c(-10, 4))  # adjust as needed
p_loadings

```

- Do the directions of the arrow align with you previous description of the PCs? (Interpretation tip: A long arrow toward PC1+ means that variable increases as you move right; toward PC2+ means it increases as you move up. Opposite directions imply negative association with that PC.)

**SOLUTION**  Yes: PC1 contrasts high in-store spend, age, income vs low; PC2 contrasts high online spend vs low.

### Plot PCA scores with clusters

```{r pca-plot}
# plot the PCA scores with clusters
ggplot(pca_data, aes(PC1, PC2, color = cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(title = "PCA Projection of Customers",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  theme_minimal() +
  theme(legend.position = "top")
```

- Do clusters separate cleanly in the first two PCs?

**SOLUTION** Yes—first two PCs (largely spend contrasts) separate the three groups visually.

### PCA Biplot (Scores + Loadings)


```{r pca-biplot}
arrow_scale <- 2   # tweak if arrows are too short/long

biplot <- ggplot() +
  geom_point(data = pca_data, aes(x = PC1, y = PC2, color = cluster), size = 2, alpha = 0.8) +
  geom_segment(data = pca_loadings,
               aes(x = 0, y = 0, xend = PC1*arrow_scale, yend = PC2*arrow_scale),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "blue") +
  geom_text(data = pca_loadings,
            aes(x = PC1 * 1.4, y = PC2 * 1.4, label = variable),
            size = 4) +
  labs(
    title = "Biplot of PCA",
    x = "\nPrincipal Component 1",
    y = "Principal Component 2\n"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  theme_minimal()
biplot
```
- How do the clusters relate to the original features based on the biplot? Do you see a correspondence with the clustering means table above (part 6)?

**SOLUTION** Yes: Cluster 1 (in-store heavy) aligns with high InStore_Spend; Cluster 2 (online heavy) aligns with high Online_Spend; Cluster 3 (omni-channel) aligns with high both.

- Now that you performed the whole clustering analysis, answer the following questions:
  - Which cluster looks most valuable? 
  - What marketing actions would you take for each segment? (e.g., how would you target them? how? what offers?)

**SOLUTION** Omni-channel (high both) likely most valuable (cross-sell, loyalty). Online-heavy: digital promotions; In-store-heavy: local offers/merchandising.



