---
title: "Customer Segmentation: Clustering and PCA"
subtitle: "MKT 566 · Fall 2026"
author: "Davide Proserpio"
institute: "USC Marshall School of Business"
format:
  revealjs:
    theme: theme.scss
    slide-number: c/t
    width: 1280
    height: 720
    incremental: false
    transition: fade
    transition-speed: fast
    controls: true
    progress: true
    hash: true
    embed-resources: false
    link-external-newwindow: true
    footer: "MKT 566 · Marketing Analytics"
execute:
  echo: false
  warning: false
  message: false
fig-width: 7
fig-height: 4.2
fig-align: center
---

```{r setup}
#| include: false
library(ggplot2)
library(data.table)
library(readxl)
library(ggrepel)
library(scales)

options(width = 110)
knitr::opts_knit$set(root.dir = "code")

# --- toy customer data for the opening example ---
set.seed(42)
toy <- rbind(
  data.table(purchases = rnorm(20, 4, 1.2),  spend = rnorm(20, 150, 40),  truth = "A"),
  data.table(purchases = rnorm(20, 18, 2),   spend = rnorm(20, 1400, 160), truth = "B"),
  data.table(purchases = rnorm(20, 16, 2),   spend = rnorm(20, 350, 60),  truth = "C")
)
km_toy <- kmeans(scale(toy[, .(purchases, spend)]), centers = 3, nstart = 20)
toy[, cluster := as.factor(km_toy$cluster)]

# --- office-store segmentation data ---
dd <- as.data.table(read_excel("code/data/segmentation_office.xlsx",
                               sheet = "SegmentationData"))
ratings <- dd[, .(variety_of_choice, electronics, furniture,
                  quality_of_service, low_prices, return_policy)]

set.seed(123)
wss <- sapply(2:10, function(k) kmeans(ratings, centers = k, nstart = 20)$tot.withinss)
elbow <- data.table(k = 2:10, wss = wss)

set.seed(123)
km <- kmeans(ratings, centers = 3, nstart = 20)
dd[, cluster := as.factor(km$cluster)]

pca <- prcomp(ratings, center = TRUE, scale. = TRUE)
scores <- data.table(pca$x[, 1:2], cluster = dd$cluster)
loadings <- as.data.table(pca$rotation[, 1:2], keep.rownames = "attribute")

seg_cols <- c("1" = "#990000", "2" = "#0072B2", "3" = "#E69F00")
seg_labs <- c("1: bargain hunters", "2: one-stop shoppers", "3: service seekers")

# --- brand perceptual-map data ---
brands <- fread("code/data/perceptual_map_office.csv")
pca_b  <- prcomp(brands[, 2:7], center = TRUE, scale. = TRUE)
scores_b   <- data.table(pca_b$x[, 1:2], brand = brands$brand)
loadings_b <- as.data.table(pca_b$rotation[, 1:2], keep.rownames = "attribute")
```

# Before We Start {.divider}

## Housekeeping

- **Airbnb case report** (`airbnb-report.html` + `ai-log.md`) due **Sunday,
  Sept. 27, 11:59 pm** on Brightspace. Tasks 6–8 are yours to finish
- **Homework 2** (box office) due **Thursday, Oct. 8**
- **Group project**: proposal presentations on **Oct. 13 and 15**
- Today: **clustering** (k-means) and **PCA**, with two scripts you run
  along with me. Tuesday: **recommendation systems**

## A different kind of question {.smaller-90}

Every model so far had an **outcome** to predict: price on Thursday, gem on
Tuesday. Today there is no outcome. There are just customers, described by
numbers, and one question: **who is similar to whom?**

::::: {.columns}
::: {.column width="50%"}
**What changes**

- No $Y$, no coefficients, no stars: the algorithm proposes **groups**
- This is called **unsupervised** learning: nothing tells the algorithm
  what the right answer is
:::
::: {.column style="width: 44%; margin-left: 5%;"}
**What stays**

- The data is still rows and columns, and **you** choose the columns
- The output still means nothing until **you** interpret it
- Still **not causation**: clusters describe, they do not explain
:::
:::::

## Why marketers cluster {.smaller-90}

Grouping similar customers is **segmentation**, and almost everything in
marketing is segment-specific:

- **Messaging**: bargain hunters get the discount email, service seekers get
  the concierge pitch
- **Product and pricing**: which features matter to which group, and who
  will pay for them
- **Budget**: which segments are growing, and where the next dollar goes

::: {.box}
Two tools today. **K-means** finds the groups. **PCA** draws the map you
need to *see* them, and it powers the classic marketing **perceptual map**
along the way.
:::

# Finding Groups: K-Means {.divider}

## Sixty customers, two numbers

```{r toy-plain}
#| fig-width: 8
#| fig-height: 4.4
p_toy <- ggplot(toy, aes(x = purchases, y = spend)) +
  geom_point(size = 3, alpha = 0.8, color = "grey35") +
  scale_y_continuous(labels = dollar) +
  labs(title = "Customers: Purchases vs. Total Spend, Last Year",
       x = "Purchases last year", y = "Total spend last year") +
  theme_minimal(base_size = 14)
p_toy
```

::: {.smaller-80 .muted}
Each dot is a customer: how often they bought, and how much they spent in
total. How many groups do you see, and how would you describe each one?
:::

## The same chart, after k-means {.smaller-90}

```{r toy-clustered}
#| fig-width: 8
#| fig-height: 3.6
cents <- toy[, .(purchases = mean(purchases), spend = mean(spend)), by = cluster]
p_toy +
  geom_point(data = toy, aes(color = cluster), size = 3, alpha = 0.85) +
  geom_point(data = cents, aes(color = cluster), shape = 4, size = 6, stroke = 2) +
  scale_color_manual(values = unname(seg_cols), guide = "none") +
  labs(title = "The Same Customers, Colored by K-Means (k = 3)")
```

- We told k-means only two things: the **numbers** and **k = 3**, the number
  of groups to find. It found the same groups your eyes did
- The **X** marks each cluster's center. The groups have obvious names:
  occasional shoppers, frequent small-basket shoppers, big spenders
- Eyes work with 2 columns and 60 customers. With **6 columns and thousands
  of customers** they do not. The algorithm scales; the logic stays the same

## How k-means works {.smaller-90}

You pick **k**, the number of clusters. The algorithm does the rest:

1. Place k **centers** at random
2. **Assign** every customer to the nearest center
3. **Move** each center to the middle of the customers assigned to it
4. Repeat 2 and 3 until the centers stop moving

- "Nearest" means plain **distance** between rows of numbers, so every
  column you feed it counts
- The random start means results can vary run to run: `set.seed()` makes
  them reproducible, and `nstart = 20` runs it 20 times and keeps the best

::: {.box}
K-means never sees names, labels, or meanings. It groups rows of numbers by
distance. Everything the clusters *mean* comes later, from you.
:::

## Watch the units: distance and scale {.smaller-90}

Suppose you cluster on **income** (\$15,000 to \$70,000) and **age** (20 to
60):

- Two customers \$1,000 apart in income are 1,000 "units" apart; 25 years
  apart in age, only 25. The distance, and so the clusters, would be
  **all income and no age**
- The fix: **standardize** first, replacing each column with z-scores
  (subtract the mean, divide by the standard deviation). In R:
  `scale(data)`. Then every column counts equally

::: {.box-warn}
Standardize whenever the columns have different units. Today's survey
ratings all share the same 1–10 scale, so we can skip it, and keep the
cluster averages in interpretable 1–10 units.
:::

## How many clusters? The elbow curve

```{r elbow-plot}
#| fig-width: 8
#| fig-height: 4.2
p_elbow <- ggplot(elbow, aes(x = k, y = wss)) +
  geom_line(color = "#990000", linewidth = 1.1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 2:10) +
  labs(title = "Total Within-Cluster Sum of Squares vs. k",
       x = "Number of clusters k", y = "Total within-cluster SS") +
  theme_minimal(base_size = 14)
p_elbow
```

::: {.smaller-80 .muted}
For each k, run k-means and add up the squared distances from every point to
its cluster center: the "spread left inside the clusters". Where does the
curve bend?
:::

## The elbow: reading it {.smaller-90}

```{r elbow-plot-2}
#| fig-width: 8
#| fig-height: 2.8
p_elbow + labs(title = NULL)
```

- Going from 2 to 3 clusters cuts the within-cluster spread from **636 to
  288**. From 3 to 4 saves only 36 more, and the curve flattens: the
  **elbow is at k = 3**
- More clusters always fit better, so "best fit" cannot pick k: with one
  cluster per customer the fit is perfect and you have learned nothing
- The elbow is a guide, not a law: k must also make **business sense**
  (three campaigns, not nine)

# Segmenting Office-Store Shoppers {.divider}

## The data: an importance survey {.smaller-90}

**40 respondents** rated how important six store attributes are to them when
buying office equipment, each on a **1–10** scale:

`variety_of_choice`, `electronics`, `furniture`, `quality_of_service`,
`low_prices`, `return_policy`

- We also know three things about each respondent: **professional** (0/1),
  **income** (in \$1,000s), and **age**
- We cluster on the **six ratings only**. The three descriptors stay out,
  and describe the segments afterwards
- The file is Excel, so a new reader: `read_excel()` from the **readxl**
  package (`fread()` is for CSVs)

::: {.smaller-80 .muted}
Data from *R for Marketing Students*. Script: `w5-1-clustering-class.R`,
data: `data/segmentation_office.xlsx`, both in `w5-code.zip`.
:::

## K-means in R: one function {.smaller}

```{r kmeans-echo}
#| echo: true
dd <- as.data.table(read_excel("data/segmentation_office.xlsx", sheet = "SegmentationData"))
ratings <- dd[, .(variety_of_choice, electronics, furniture,
                  quality_of_service, low_prices, return_policy)]
set.seed(123)                                     # random starts, reproducible
km <- kmeans(ratings, centers = 3, nstart = 20)   # k = 3, keep the best of 20 runs
km$size                                           # respondents per cluster
```

```{r attach-cluster}
#| echo: true
dd[, cluster := as.factor(km$cluster)]            # attach the labels: 1, 2, or 3
```

- Three clusters of **14, 18, and 8** respondents. Each respondent now
  carries a cluster label
- The labels are just 1, 2, 3, in no particular order. What they *mean*
  is the next slide's job

## The profile table {.smaller}

The average rating of each attribute, within each cluster:

```{r profiles-echo}
#| echo: true
dd[, lapply(.SD, function(x) round(mean(x), 1)), by = cluster,
   .SDcols = names(ratings)][order(cluster)]
```

::: {.smaller-80 .muted}
Read it row by row: what does each cluster rate **high**, and what does it
rate **low**? Try to give each row a name before the next slide.
:::

## Naming the segments {.smaller-90}

| Cluster (n) | Rates high | Rates low | The segment |
|---|---|---|---|
| 1 (n = 14) | low prices **8.3**, return policy **6.3** | furniture 1.4, electronics 2.8 | **bargain hunters** |
| 2 (n = 18) | variety **9.1**, electronics **6.1**, furniture **5.8** | service 2.4, low prices 3.7 | **one-stop shoppers** |
| 3 (n = 8) | quality of service **8.5** | low prices 2.5, furniture 1.8 | **service seekers** |

- The names come from the **table**, not from the algorithm: k-means only
  ever says 1, 2, 3
- Check the names against the low ratings too: one-stop shoppers do not
  care about price; service seekers actively dislike discount-store vibes

::: {.box}
A segment name is a claim about the table. If a colleague cannot trace your
name back to the averages, rename it.
:::

## Who is in each segment? {.smaller-90}

The three columns we kept **out** of the clustering describe who landed in
each segment:

| Segment | Professional | Income | Age |
|---|---|---|---|
| Bargain hunters (14) | 50% | \$32,000 | 31 |
| One-stop shoppers (18) | 33% | \$48,000 | 44 |
| Service seekers (8) | **75%** | \$48,000 | **49** |

- Bargain hunters are the **young, lower-income** group; service seekers are
  **older professionals**; one-stop shoppers sit in between with the largest
  headcount
- This is the actionable half: ratings say **what to offer** each segment,
  descriptors say **who and where they are**

::: {.box}
Because these columns played no part in forming the clusters, the
differences across segments are a finding, not something we built in.
:::

# Six Dimensions on One Screen: PCA {.divider}

## The problem, and the squash {.smaller-90}

The toy chart worked because there were **two** columns. Our shoppers live
in **six**. A screen has two axes; you cannot plot six.

- **PCA** (principal component analysis) squashes the six columns into
  **two new axes**, chosen so the map keeps as much of the respondents'
  **variation** as possible (Tuesday we squashed a score into a probability;
  today we squash six columns into two axes)
- Each respondent gets a coordinate on each new axis: **PC1** and **PC2**.
  Plot those two and you have the best two-axis picture of the data

```{r pca-echo}
#| echo: true
pca <- prcomp(ratings, center = TRUE, scale. = TRUE)
round(summary(pca)$importance[2, ], 2)   # share of variation per component
```

- PC1 keeps **52%** of the variation, PC2 another **26%**: the 2D map shows
  **78%** of what six columns contain. Good enough to look at

## What do the two axes mean? {.smaller-90}

::::: {.columns}
::: {.column width="55%"}
```{r loadings-plot}
#| fig-width: 6.2
#| fig-height: 4.8
arrow_scale <- 3
ggplot(loadings, aes(x = PC1 * arrow_scale, y = PC2 * arrow_scale, label = attribute)) +
  geom_vline(xintercept = 0, linetype = 3, color = "grey60") +
  geom_hline(yintercept = 0, linetype = 3, color = "grey60") +
  geom_segment(aes(x = 0, y = 0, xend = PC1 * arrow_scale, yend = PC2 * arrow_scale),
               arrow = arrow(length = unit(0.2, "cm")), color = "#990000") +
  geom_text_repel(size = 4.5) +
  labs(title = "PCA loadings: the six ratings as arrows", x = "PC1", y = "PC2") +
  theme_minimal(base_size = 13)
```
:::
::: {.column width="45%"}
Each arrow is one of the six ratings; its direction shows where that rating
**pulls** a respondent on the map:

- **Left**: variety, electronics, furniture. The left side of the map is
  "wants a big assortment"
- **Upper right**: low prices, return policy. **Lower right**: quality of
  service
- So PC1 separates *assortment* from *price and service*; PC2 separates
  *price* from *service*
:::
:::::

::: {.smaller-80 .muted}
The signs are arbitrary (R could have flipped left and right). Directions
and groupings are what matter.
:::

## The map: 40 shoppers, colored by cluster

```{r cluster-map}
#| fig-width: 8
#| fig-height: 4.4
p_map <- ggplot(scores, aes(x = PC1, y = PC2, color = cluster)) +
  geom_vline(xintercept = 0, linetype = 3, color = "grey60") +
  geom_hline(yintercept = 0, linetype = 3, color = "grey60") +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = seg_cols, labels = seg_labs, name = NULL) +
  labs(title = "Respondents on the PCA Map, Colored by K-Means Cluster",
       x = "PC1", y = "PC2") +
  theme_minimal(base_size = 14) + theme(legend.position = "top")
p_map
```

::: {.smaller-80 .muted}
PCA drew the map; k-means picked the colors. The two never talked to each
other. Do the colors form regions?
:::

## The map: reading it {.smaller-90}

```{r biplot}
#| fig-width: 8
#| fig-height: 3.4
arrow_scale <- 3
p_map +
  geom_segment(data = loadings,
               aes(x = 0, y = 0, xend = PC1 * arrow_scale, yend = PC2 * arrow_scale),
               inherit.aes = FALSE,
               arrow = arrow(length = unit(0.2, "cm")), color = "grey40") +
  geom_text_repel(data = loadings,
                  aes(x = PC1 * arrow_scale, y = PC2 * arrow_scale, label = attribute),
                  inherit.aes = FALSE, size = 3.8, color = "grey30") +
  labs(title = "The Same Map, with the Attribute Arrows (a Biplot)")
```

- The clusters occupy **separate regions**, and each sits at the tip of the
  arrows its profile predicted: bargain hunters by *low prices*, one-stop
  shoppers by *variety and furniture*, service seekers by *quality of
  service*
- Two independent tools agree: the segments are in the data, not an
  artifact of one algorithm

## The same trick on brands: a perceptual map

```{r brand-map}
#| fig-width: 8.5
#| fig-height: 4.4
arrow_scale_b <- 2
ggplot() +
  geom_vline(xintercept = 0, linetype = 3, color = "grey60") +
  geom_hline(yintercept = 0, linetype = 3, color = "grey60") +
  geom_segment(data = loadings_b,
               aes(x = 0, y = 0, xend = PC1 * arrow_scale_b, yend = PC2 * arrow_scale_b),
               arrow = arrow(length = unit(0.2, "cm")), color = "grey40") +
  geom_text_repel(data = loadings_b,
                  aes(x = PC1 * arrow_scale_b, y = PC2 * arrow_scale_b, label = attribute),
                  size = 4, color = "grey30") +
  geom_point(data = scores_b, aes(x = PC1, y = PC2), color = "#990000", size = 4) +
  geom_text_repel(data = scores_b, aes(x = PC1, y = PC2, label = brand),
                  size = 5, fontface = "bold", color = "#990000") +
  labs(title = "Perceptual Map: Four Office-Equipment Brands, Six Attributes",
       x = "PC1", y = "PC2") +
  theme_minimal(base_size = 14)
```

::: {.smaller-80 .muted}
Same PCA, new rows: **brands** instead of shoppers, rated on six attributes.
Which brands compete head-to-head? Who owns which attribute?
:::

## The perceptual map: reading it {.smaller-90}

- **OfficeStar** sits alone on the right, with *service, product quality,
  large choice*, and, crucially, the **preference** arrow: the strong brand
  people like best
- **Supermarket** is its mirror image: far left, near *low prices* and
  *convenience*, far from quality and preference. **PaperNCo** owns the
  cheap corner; **OfficeEquipment** competes on convenience with no clear
  strength
- Two axes hold **98%** of the variation here, so this map is essentially
  the whole story

::: {.box}
Perceptual maps answer positioning questions: who are my real competitors
(neighbors on the map), what do I own (my nearest arrows), and where is the
**gap** nobody occupies. Script: `w5-1-pca-class.R`.
:::

## Clustering and PCA, side by side {.smaller-90}

| | K-means | PCA |
|---|---|---|
| Question | who is similar to whom? | which directions matter? |
| Output | a **group label** per row (1, 2, 3) | **two new columns** (PC1, PC2) per row |
| You choose | the columns, and k | the columns |
| Read it via | the profile table | the loadings (arrows) |
| Marketing use | segmentation | perceptual maps, visualization |
| Together | PCA draws the map, k-means draws the borders | |

::: {.box-warn}
Cautions that apply to both: standardize when units differ; the labels and
axes mean nothing until you read them off a table; and clusters
**describe** customers, they do not explain them. "Service seekers are
older" is a pattern, not a mechanism.
:::

## Wrap-up

::: {.box}
Code and data to reproduce everything, on the course website:

- [**`w5-code.zip`**](https://github.com/dadepro/mkt566/raw/main/w5/w5-code.zip):
  both scripts and both datasets
- [**`w5-1-clustering-class.R`**](https://github.com/dadepro/mkt566/blob/main/w5/code/w5-1-clustering-class.R):
  elbow, k-means, profiles, the PCA map
- [**`w5-1-pca-class.R`**](https://github.com/dadepro/mkt566/blob/main/w5/code/w5-1-pca-class.R):
  the brand perceptual map
:::

::: {.smaller-80 .muted}
Required reading: chapters
[5](https://bookdown.org/content/6ef13ea6-4e86-4566-b665-ebcd19d45029/pca_office.html),
[6](https://bookdown.org/content/6ef13ea6-4e86-4566-b665-ebcd19d45029/pca_toothpase.html),
and
[7](https://bookdown.org/content/6ef13ea6-4e86-4566-b665-ebcd19d45029/cluster.html)
of *R for Marketing Students*. Optional: chapters 11.1–11.3 of *R for
Marketing Research and Analytics*. Tuesday: recommendation systems.
:::

# Questions? {.divider}
