---
title: "Regression: Modeling Relationships"
subtitle: "MKT 566 · Fall 2026"
author: "Davide Proserpio"
institute: "USC Marshall School of Business"
format:
  revealjs:
    theme: theme.scss
    slide-number: c/t
    width: 1280
    height: 720
    incremental: false
    transition: fade
    transition-speed: fast
    controls: true
    progress: true
    hash: true
    embed-resources: false
    link-external-newwindow: true
    footer: "MKT 566 · Marketing Analytics"
execute:
  echo: false
  warning: false
  message: false
fig-width: 7
fig-height: 4.2
fig-align: center
---

```{r setup}
#| include: false
library(datarium)
library(ggplot2)
library(data.table)
library(patchwork)
library(scales)
library(stargazer)

options(width = 110, scipen = 10)
set.seed(566)
knitr::opts_knit$set(root.dir = "code")

data("marketing", package = "datarium")
# The package stores spend in thousands of dollars and sales in thousands of
# units. Put both in natural units (dollars, units) so every coefficient
# reads "per $1".
marketing[] <- lapply(marketing, function(v) v * 1000)
airbnb <- fread("code/data/airbnb.csv")

m_yt   <- lm(sales ~ youtube, data = marketing)
b_yt   <- coef(m_yt)
ssr_yt <- sum(resid(m_yt)^2)
sst    <- sum((marketing$sales - mean(marketing$sales))^2)
```

# Before We Start {.divider}

## Housekeeping

- The **RateBeer report** (`ratebeer-report.html` + `ai-log.md`) is due
  **Sunday, Sept. 20, 11:59 pm** on Brightspace
- **Homework 2** is released today and due **Thursday, Oct. 8**. It covers
  today (OLS) and Tuesday (logit), on a box-office dataset
- **Group project**: proposal presentations are on **Oct. 13 and 15**
  (10–15 minutes per group: the problem you will study and how)
- Tuesday: **logit**, then the **Airbnb exercise** on your laptop

## From charts to models {.smaller-90}

Last week you *looked* at how two variables move together. Today you put a
**number** on it, and learn to tell when that number can be trusted.

::::: {.columns}
::: {.column width="45%"}
**Today you learn to**

1. Say what a regression **is** (a line, chosen a specific way)
2. **Read** R's output: coefficient, standard error, p-value, R²
3. **Interpret** a coefficient in business units, with and without logs
4. Handle **categorical** predictors (city, room type)
:::
::: {.column width="45%" style="margin-left: 5%;"}
**What stays the same**

- The AI writes the code. You choose the model, and you read the result
- Two habits: **predict** before you run, **explain back** one line after
- Same data as before: the 200 ad campaigns from week 3, then 50,000
  Airbnb listings
:::
:::::

::: {.box}
Regression is the workhorse of marketing analytics. Homework 2, the Airbnb
exercise, and most group projects are built on what we cover today.
:::

## R script for today

In the `code/` folder on the course website:

1. **`w4-1-ols-class.R`**: every chart and regression from the first half
   (the marketing data), with a comment above each one
2. **`w4-1-airbnb-class.R`**: the second half, on the Airbnb data
3. **`data/airbnb.csv`**: the Airbnb dataset (also used in the Airbnb
   exercise on Tuesday, Sept. 22)

::: {.box}
Download everything as one zip:
[**`w4-code.zip`**](https://github.com/dadepro/mkt566/raw/main/w4/w4-code.zip),
open the folder in VS Code, and run along.
:::

# From a Line to a Model {.divider}

## The line we drew last week

```{r line-recall}
#| echo: true
#| output-location: column
#| code-line-numbers: "3"
#| fig-width: 5.5
#| fig-height: 4.2
ggplot(marketing, aes(x = youtube, y = sales)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Sales vs. YouTube Ad Spend",
       x = "YouTube ad spend ($)",
       y = "Sales (units)") +
  theme_minimal()
```

::: {.smaller-80 .muted}
We asked R for a straight line through the points. What line is it? What
could a manager do with it?
:::

## A regression is the formula for that line

$$\text{sales} = \beta_0 + \beta_1 \times \text{youtube}$$

- A regression **predicts** one variable, $Y$, from one or more others, $X$
- $Y$ is the **outcome** (or dependent variable). $X$ is the **predictor**
  (or independent variable)
- Examples:
  - Sales ($Y$) as a function of ad spend ($X$)
  - Revenue ($Y$) as a function of review ratings ($X$)
  - Airbnb price ($Y$) as a function of the number of bedrooms ($X$)
  - Income ($Y$) as a function of years of education ($X$)

::: {.box}
In general $Y = F(X)$: $Y$ depends on $X$ *somehow*. A **linear** regression
assumes the simplest $F$: a straight line. Machine learning (week 10) is
mostly about learning fancier $F$'s.
:::

## The setup

::::: {.columns}
::: {.column width="60%"}
For each row $i$ of the data (each campaign):

$$y_i = \beta_0 + \beta_1 x_i + \epsilon_i$$

- $x_i$: the predictor (YouTube spend of campaign $i$)
- $y_i$: the outcome (sales of campaign $i$)
- $\beta_0$: the **intercept**, predicted $Y$ when $X = 0$
- $\beta_1$: the **slope**, how much $Y$ changes when $X$ goes up by one
  unit
- $\epsilon_i$: the **error**, the part of $y_i$ the line does not explain
:::
::: {.column width="40%"}
```{r setup-table}
#| echo: false
knitr::kable(head(marketing[, c("youtube", "sales")], 5),
             col.names = c("x = youtube ($)", "y = sales (units)"),
             digits = 0, format.args = list(big.mark = ","))
```
::: {.smaller-80 .muted}
Rows $i = 1, \dots, 200$
:::
:::
:::::

## The three pieces, on the chart

```{r ols-anatomy}
#| fig-width: 8
#| fig-height: 4.6
pt <- marketing[which.max(resid(m_yt)), ]
pt_hat <- predict(m_yt, pt)
x1 <- 270000; x2 <- 340000
y1 <- b_yt[1] + b_yt[2] * x1; y2 <- b_yt[1] + b_yt[2] * x2
ggplot(marketing, aes(x = youtube, y = sales)) +
  geom_point(alpha = 0.3, color = "darkgreen") +
  geom_abline(intercept = b_yt[1], slope = b_yt[2], color = "firebrick", linewidth = 1.1) +
  annotate("point", x = 0, y = b_yt[1], size = 3.5, color = "firebrick") +
  annotate("label", x = 12000, y = b_yt[1] - 3200, hjust = 0, parse = TRUE,
           label = sprintf("beta[0] == %.0f ~~ (intercept)", b_yt[1]), color = "firebrick") +
  annotate("segment", x = x1, xend = x2, y = y1, yend = y1, color = "firebrick", linetype = 2) +
  annotate("segment", x = x2, xend = x2, y = y1, yend = y2, color = "firebrick", linetype = 2) +
  annotate("label", x = x2, y = y1 - 2200, hjust = 1, parse = TRUE,
           label = sprintf("beta[1] == %.3f ~~ (slope)", b_yt[2]), color = "firebrick") +
  annotate("segment", x = pt$youtube, xend = pt$youtube, y = pt_hat, yend = pt$sales,
           color = "steelblue", linewidth = 1.1) +
  annotate("point", x = pt$youtube, y = pt$sales, size = 3, color = "steelblue") +
  annotate("label", x = pt$youtube - 8000, y = (pt$sales + pt_hat) / 2, hjust = 1, parse = TRUE,
           label = "epsilon == actual - predicted ~~ (error)", color = "steelblue") +
  scale_x_continuous(limits = c(0, 360000), labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(x = "YouTube ad spend ($)", y = "Sales (units)") +
  theme_minimal(base_size = 14)
```

::: {.smaller-80 .muted}
Every point has its own error: the vertical distance to the line. The
intercept and the slope are the same for everyone.
:::

## Which line?

```{r which-line}
#| fig-width: 10
#| fig-height: 4.2
draw_line <- function(a, b, title) {
  d <- copy(as.data.table(marketing))[, yhat := a + b * youtube]
  ggplot(d, aes(x = youtube, y = sales)) +
    geom_segment(aes(xend = youtube, yend = yhat), color = "steelblue", alpha = 0.5) +
    geom_point(alpha = 0.4, color = "darkgreen") +
    geom_abline(intercept = a, slope = b, color = "firebrick", linewidth = 1.1) +
    scale_x_continuous(labels = comma) +
    scale_y_continuous(labels = comma) +
    labs(title = title, x = "YouTube ad spend ($)", y = "Sales (units)") +
    theme_minimal(base_size = 13)
}
draw_line(5000, 0.08, "Line A") + draw_line(b_yt[1], b_yt[2], "Line B")
```

::: {.smaller-80 .muted}
Both lines run through the points. The blue segments are the errors. Which
line is better, and how would you *decide*?
:::

## Which line: reading it

```{r which-line-2}
#| fig-width: 10
#| fig-height: 3.6
ssr_a <- sum((marketing$sales - (5000 + 0.08 * marketing$youtube))^2)
draw_line(5000, 0.08, sprintf("Line A\nsum of squared errors = %s", comma(round(ssr_a)))) +
  draw_line(b_yt[1], b_yt[2], sprintf("Line B\nsum of squared errors = %s", comma(round(ssr_yt))))
```

::: {.smaller-90}
- Square each error (so misses above and below the line both count), add
  them up. Smaller is better
- **Ordinary least squares (OLS)** picks the intercept and slope with the
  **smallest** sum of squared errors. Line B is that line, and it is what
  `geom_smooth(method = "lm")` drew last week
- "Least squares" is the whole idea. Everything else today is about reading
  what comes out
:::

## The sum of squared errors, written out {.smaller-90}

::::: {.columns}
::: {.column width="58%"}
For any intercept $\beta_0$ and slope $\beta_1$, the line predicts
$\hat{y}_i = \beta_0 + \beta_1 x_i$ and misses by $\epsilon_i = y_i - \hat{y}_i$:

$$\text{SSE}(\beta_0, \beta_1) = \sum_{i=1}^{n} \epsilon_i^2
= \sum_{i=1}^{n} \left(y_i - \beta_0 - \beta_1 x_i\right)^2$$

- Every choice of $(\beta_0, \beta_1)$ gives a different line, hence a
  different SSE. Line A and Line B are two such choices
- OLS picks the pair that makes this sum as small as possible:

$$(\hat{\beta}_0, \hat{\beta}_1) = \arg\min_{\beta_0,\, \beta_1} \; \text{SSE}(\beta_0, \beta_1)$$
:::
::: {.column width="42%"}
```{r sse-table}
#| echo: false
sse_tab <- data.table(
  x = marketing$youtube[1:4], y = marketing$sales[1:4],
  yhat = fitted(m_yt)[1:4], e = resid(m_yt)[1:4])[, e2 := e^2]
sse_tab <- rbind(sse_tab, data.table(x = NA, y = NA, yhat = NA, e = NA, e2 = NA),
                 data.table(x = NA, y = NA, yhat = NA, e = NA, e2 = ssr_yt))
knitr::kable(sse_tab, digits = 0, format.args = list(big.mark = ","),
             col.names = c("$x_i$", "$y_i$", "$\\hat{y}_i$", "$\\epsilon_i$", "$\\epsilon_i^2$"),
             row.names = FALSE) |>
  gsub(pattern = "NA", replacement = "⋮")
```
::: {.smaller-80 .muted}
Line B (the OLS line), first 4 of 200 campaigns. Bottom row: the sum over
all 200, the number on the previous slide.
:::
:::
:::::

::: {.box}
"Least squares" means exactly this: square each error, add them up, and pick
the line that makes the total smallest.
:::

# Estimating a Regression in R {.divider}

## One function: `lm()` {.smaller}

```{r lm-first}
#| echo: true
model <- lm(sales ~ youtube, data = marketing)   # "sales as a function of youtube"
summary(model)
```

::: {.smaller-80 .muted}
Find the intercept and the slope in this output. What else is in there?
:::

## Reading the output: the estimates {.smaller-90}

```{r lm-coef}
#| echo: true
coef(model)
```

- **Intercept** $\beta_0 = 8{,}439$: a campaign with **no** YouTube spend is
  predicted to sell about **8,400 units**
- **Slope** $\beta_1 = 0.0475$: each additional **\$1** on YouTube is
  associated with **0.0475 more units** sold
- Prediction: with \$100,000 on YouTube,
  $8{,}439 + 0.0475 \times 100{,}000 \approx 13{,}200$ units

::: {.box}
Always translate the slope into **business units**: "one more unit of $X$
goes with $\beta_1$ more units of $Y$". If you cannot say it in a sentence a
manager understands, you have not read the coefficient yet.
:::

## The estimate comes with uncertainty {.smaller-90}

Our slope comes from **these** 200 campaigns. Another 200 would give a
slightly different slope. The **standard error (SE)** measures how much it
would move around.

```{r lm-se}
#| echo: true
round(coef(summary(model)), 4)
```

- `Std. Error` of the slope: **0.0027**. The slope is pinned down to roughly
  $0.0475 \pm 0.005$
- `t value` = estimate / SE = $0.0475 / 0.0027 \approx$ **17.7**: the slope is
  about 18 standard errors away from zero
- `Pr(>|t|)` is the **p-value**: here so small that it rounds to **0**
  (two slides back, `summary()` printed it as `<2e-16`)

## The question the p-value answers {.smaller-90}

::::: {.columns}
::: {.column width="45%"}
**Hypothesis testing** is a way to decide between two claims with data:

- $H_0$ (the null): **no relationship**, $\beta_1 = 0$. YouTube spend has
  nothing to do with sales
- $H_1$ (the alternative): **there is one**, $\beta_1 \neq 0$

The **p-value** answers: *if $H_0$ were true, how likely is a slope at least
this far from zero, just by luck?*
:::
::: {.column width="45%" style="margin-left: 5%;"}
**Decision rule**

- p-value **< 0.05**: reject $H_0$. The coefficient is **statistically
  significant** (different from zero)
- p-value **≥ 0.05**: fail to reject $H_0$. Not enough evidence

R's stars: `***` p < 0.001, `**` p < 0.01, `*` p < 0.05, `.` p < 0.1

**Smaller SE → larger t → smaller p → stronger evidence** against "no
relationship"
:::
:::::

::: {.box}
For YouTube, p is essentially 0: if spend and sales were unrelated, a slope
of 0.0475 would almost never show up. So we are confident the relationship
is real.
:::

## "Significant" is not the same as "big" {.smaller-90}

Two separate questions about every coefficient:

1. **Is it different from zero?** That is what the p-value answers. It
   depends on the size of the estimate *and* on how much data you have
2. **Does it matter?** That is about the **size** of the estimate, in
   business units. The p-value says nothing about it

::: {.box-warn}
With 50,000 rows (Airbnb, later today) almost everything is "significant",
including effects too small to care about. With 18 rows almost nothing is,
including effects that are real. Read the **size** first, then the stars.
:::

## How good is the line? R-squared

```{r r2-figure}
#| fig-width: 10
#| fig-height: 3.9
sub <- as.data.table(marketing)[sample(.N, 40)]
sub[, `:=`(ybar = mean(marketing$sales), yhat = predict(m_yt, sub))]
p_mean <- ggplot(sub, aes(youtube, sales)) +
  geom_segment(aes(xend = youtube, yend = ybar), color = "steelblue", alpha = 0.6) +
  geom_hline(yintercept = mean(marketing$sales), color = "grey30", linewidth = 1) +
  geom_point(color = "darkgreen") +
  scale_x_continuous(labels = comma) + scale_y_continuous(labels = comma) +
  labs(title = "Errors if we only knew the mean", x = "YouTube spend ($)", y = "Sales (units)") +
  theme_minimal(base_size = 13)
p_line <- ggplot(sub, aes(youtube, sales)) +
  geom_segment(aes(xend = youtube, yend = yhat), color = "steelblue", alpha = 0.6) +
  geom_abline(intercept = b_yt[1], slope = b_yt[2], color = "firebrick", linewidth = 1) +
  geom_point(color = "darkgreen") +
  scale_x_continuous(labels = comma) + scale_y_continuous(labels = comma) +
  labs(title = "Errors with the regression line", x = "YouTube spend ($)", y = NULL) +
  theme_minimal(base_size = 13)
p_mean + p_line
```

::: {.smaller-80 .muted}
Forty of the 200 campaigns. Without $X$, the best guess for every campaign
is the average. How much do the errors shrink once we use the line?
:::

## R-squared: reading it {.smaller-90}

$$R^2 = 1 - \frac{\text{sum of squared errors with the line}}{\text{sum of squared errors with the mean}}
= 1 - \frac{`r comma(round(ssr_yt))`}{`r comma(round(sst))`} = `r round(1 - ssr_yt / sst, 2)`$$

- **Multiple R-squared: 0.61**. YouTube spend explains **61%** of the
  variation in sales across campaigns. The other 39% is other channels,
  seasonality, luck
- $R^2$ runs from 0 (the line is no better than the mean) to 1 (every point
  on the line)
- **Residual standard error: 3,910**. The typical prediction misses by
  about **3,900 units**. Same idea as $R^2$, but in the units of $Y$
- **Adjusted R-squared** is $R^2$ with a small penalty for every predictor
  you add. Use it to compare models with different numbers of predictors

::: {.box}
A low $R^2$ does not make a coefficient wrong. It says that $X$ alone leaves
most of $Y$ unexplained. Whether that matters depends on your goal:
**prediction** needs a high $R^2$, **understanding one relationship** does
not.
:::

## The output, decoded {.smaller-90}

| In the output | What it is | How to read it |
|---|---|---|
| `Estimate` | $\hat\beta$, the coefficient | "One more unit of $X$ goes with $\hat\beta$ more units of $Y$" |
| `Std. Error` | Precision of the estimate | Smaller is better. Would the estimate move much with new data? |
| `t value` | Estimate / SE | How many SEs from zero. Beyond ±2 is "significant" |
| `Pr(>|t|)` | p-value | Chance of a t this large if the true coefficient is 0 |
| `Residual standard error` | Typical size of the error | In units of $Y$ |
| `R-squared` | Share of the variation in $Y$ explained | 0 to 1 |
| `F-statistic` | Do the predictors explain *anything* together? | Tiny p-value: yes |

# Interpreting the Coefficient {.divider}

## What does $\beta_1$ tell us? {.smaller-90}

**All else equal** (*ceteris paribus*), how much $Y$ changes when $X$ moves
by one unit. What "one unit" means depends on how you wrote the model:

| Model | Equation | Reading of $\beta_1$ |
|---|---|---|
| Level-level | $Y = \beta_0 + \beta_1 X$ | One more unit of $X$ → $\beta_1$ more **units** of $Y$ |
| Log-level | $\ln Y = \beta_0 + \beta_1 X$ | One more unit of $X$ → about $100 \times \beta_1$ **percent** more $Y$ |
| Level-log | $Y = \beta_0 + \beta_1 \ln X$ | 1% more $X$ → about $\beta_1 / 100$ more **units** of $Y$ |
| Log-log | $\ln Y = \beta_0 + \beta_1 \ln X$ | 1% more $X$ → about $\beta_1$ **percent** more $Y$ (an **elasticity**) |

::: {.smaller-80 .muted}
The percentage readings are approximations that work when $\beta_1$ is
small (say below 0.2). The exact formula for log-level is
$(e^{\beta_1} - 1) \times 100\%$.
:::

## Why take logs? {.smaller-90}

::::: {.columns}
::: {.column width="50%"}
**1. Skewed variables.** Price, income, sales, number of reviews: many small
values, a few huge ones. You saw in week 2 that a log scale makes such
distributions readable. The same is true for a regression: without the log,
the few huge values pull the line around.

**2. Multiplicative relationships.** A \$1 price increase means something
very different at \$5 and at \$100. In logs, the model works in
**percentages**, which is how the relationship usually behaves.
:::
::: {.column width="50%"}
**3. Interpretation.** Managers think in percentages: "a 10% price increase
lowers demand by 8%". A log-log coefficient **is** that number, the
**elasticity**.

**Rule of thumb**: log a variable that is positive and right-skewed (money,
counts). Leave ratings, shares, and dummies as they are.
:::
:::::

::: {.box}
Logs change the **question** you are asking, not just the fit. Level: "how
many more units?". Log: "what percent more?". Choose the one your manager
needs.
:::

## The same relationship, three ways {.smaller}

```{r three-forms}
#| echo: true
m_level  <- lm(sales ~ youtube, data = marketing)
m_loglev <- lm(log(sales) ~ youtube, data = marketing)
m_loglog <- lm(log(sales) ~ log(youtube), data = marketing)
stargazer(m_level, m_loglev, m_loglog, type = "text",
          column.labels = c("level-level", "log-level", "log-log"),
          omit.stat = c("f", "ser", "adj.rsq"), digits = 4, digits.extra = 4,
          no.space = TRUE)
```

## Three ways: reading it {.smaller-90}

- **Level-level** (column 1): \$1 more on YouTube → **0.0475 more units**
- **Log-level** (column 2): \$1 more on YouTube → about **0.0003% more
  sales** ($100 \times 0.000003$)
- **Log-log** (column 3): 1% more YouTube spend → **0.36% more sales**. That
  is the **elasticity** of sales with respect to YouTube spend

::: {.box}
`stargazer()` puts several models side by side, with standard errors in
parentheses and stars for significance. This is the table format for
Homework 2 and for your project. Column labels, a title, and readable
variable names are the difference between a table and a wall of numbers.
:::

::: {.smaller-80 .muted}
Elasticities below 1 mean $Y$ responds less than proportionally: doubling
YouTube spend does not double sales.
:::

## More than one predictor {.smaller-90}

$$y_i = \beta_0 + \beta_1 x_{1,i} + \beta_2 x_{2,i} + \dots + \beta_k x_{k,i} + \epsilon_i$$

Everything we said still applies, with one change in the reading: each
$\beta_j$ is the change in $Y$ for one more unit of $x_j$, **holding the
other predictors fixed**.

In R, add predictors to the formula with `+`:

```{r multiple-code}
#| echo: true
#| eval: false
m_all <- lm(sales ~ youtube + facebook + newspaper, data = marketing)
```

::: {.box}
Before you look: last week's charts showed YouTube strong, Facebook noisier,
newspaper flat. Which coefficient do you expect to be largest? Which one
close to zero?
:::

## Three channels at once {.smaller}

```{r multiple}
#| echo: true
m_all <- lm(sales ~ youtube + facebook + newspaper, data = marketing)
stargazer(m_yt, m_all, type = "text", column.labels = c("YouTube only", "all three"),
          omit.stat = c("f", "ser", "adj.rsq"), digits = 3, no.space = TRUE)
```

::: {.smaller-80 .muted}
Which channel has the largest coefficient? Which one is essentially zero?
:::

## More than one predictor: reading it {.smaller-90}

- **Facebook** has the largest coefficient: \$1 more on Facebook,
  holding YouTube and newspaper fixed, goes with about **0.19 more units**.
  YouTube: 0.046 units per \$1
- **Newspaper**: $-0.001$, p-value 0.86. Once YouTube and Facebook are in
  the model, newspaper spend tells us **nothing** about sales. Remember
  last week's flat newspaper panel
- $R^2$ goes from 0.61 to **0.90**: three channels explain 90% of the
  variation in sales

::: {.box}
"Holding the others fixed" is what makes multiple regression useful.
Campaigns that spend more on newspaper also spend more on Facebook
(correlation 0.35). The single-predictor regression of sales on newspaper
would give newspaper credit for Facebook's work.
:::

# The Airbnb Data {.divider}

## The dataset {.smaller}

```{r airbnb-head}
#| echo: true
airbnb <- fread("data/airbnb.csv")
names(airbnb)
head(airbnb[, .(price, bedrooms, star_rating, reviews_count, city, room_type)], 4)  # a few columns
table(airbnb$city)
```

::: {.smaller-90}
`r comma(nrow(airbnb))` listings in four cities, with a nightly price,
size, star rating, number of reviews, city, and room type.
:::

## Predict before you run

Let's predict **price** as a function of the **number of reviews** a listing
has.

::: {.box}
Write down one sentence: **what sign** do you expect for the slope, and
**why**?
:::

::: {.smaller-90}
Two stories, opposite signs:

- More reviews means a popular, established listing, which can charge more
  (**positive**)
- Cheap places get booked more often, so they collect more reviews
  (**negative**)
:::

## Price vs. number of reviews

```{r airbnb-scatter}
#| echo: true
#| output-location: column
#| fig-width: 5.5
#| fig-height: 4.2
ggplot(airbnb, aes(x = reviews_count, y = price)) +
  geom_point(alpha = 0.2) +
  scale_y_continuous(labels = dollar) +
  labs(title = "Price vs. Number of Reviews",
       x = "Number of reviews",
       y = "Nightly price") +
  theme_minimal()
```

::: {.smaller-80 .muted}
`r comma(nrow(airbnb))` points. What do you see? Does a straight line make
sense here?
:::

## Price vs. reviews: reading it {.nostretch}

```{r airbnb-scatter-2}
#| fig-width: 6
#| fig-height: 3.1
ggplot(airbnb, aes(x = reviews_count, y = price)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  scale_y_continuous(labels = dollar) +
  labs(title = "Price vs. Number of Reviews", x = "Number of reviews", y = "Nightly price") +
  theme_minimal()
```

::: {.smaller-90}
- Both variables are **heavily skewed**: median price \$100, a few listings
  at \$10,000; median 8 reviews, a few above 300
- No clear line. If anything, the pricey listings sit at **low** review
  counts, so the line tilts slightly **down**
- A prime candidate for logs (the exercise on Tuesday, Sept. 22). For now,
  let's see what
  the straight line says
:::

## The regression {.smaller}

```{r airbnb-lm}
#| echo: true
m1 <- lm(price ~ reviews_count, data = airbnb)
summary(m1)
```

::: {.smaller-80 .muted}
Read the slope, the p-value, and the R-squared. Do they agree with the
scatter plot?
:::

## The regression: reading it {.smaller-90}

- **Intercept**: a listing with zero reviews is predicted to cost **\$148**
- **Slope** $-0.35$: each additional review is associated with a price
  **35 cents lower**. Ten more reviews: \$3.50 lower. That is tiny next to a
  median price of \$100
- **p-value < 2e-16**, three stars: statistically significant. With
  50,000 listings, even a 35-cent slope is measured precisely
- **R-squared 0.002**: reviews explain **0.2%** of the variation in price
- **Residual standard error \$166**: the typical miss is larger than the
  typical price

::: {.box-warn}
Significant, and useless for prediction. The stars say the slope is not
zero. The size and the $R^2$ say it does not matter. Both readings are
correct, and you need both.
:::

## A table you can put in a report {.smaller}

```{r airbnb-stargazer}
#| echo: true
stargazer(m1, type = "text",
          title = "Regression of Price on Number of Reviews",
          dep.var.labels = "Nightly price (dollars)",
          covariate.labels = "Number of reviews",
          omit.stat = c("f", "ser", "adj.rsq"), digits = 2, no.space = TRUE)
```

::: {.smaller-80 .muted}
Same numbers as `summary()`, laid out for a reader. Standard errors in
parentheses, stars for significance, $N$ and $R^2$ at the bottom.
:::

# Categorical Predictors {.divider}

## Text in a regression? {.smaller-90}

City is a word, not a number. R handles this with a **factor**: a
categorical variable stored as numbers with labels attached.

```{r factor-demo}
#| echo: true
city <- factor(c("Miami", "Austin", "Miami", "Boston"))
city
as.integer(city)
```

- R stores 1, 2, 3 and shows you the labels. Levels are sorted
  alphabetically unless you say otherwise
- `lm()` converts a text column into a factor automatically. But **what does
  a slope on "city" mean?** Let's look

## Price by city {.smaller}

```{r airbnb-city}
#| echo: true
m_city <- lm(price ~ city, data = airbnb)
stargazer(m_city, type = "text", title = "Regression of Price on City",
          dep.var.labels = "Nightly price (dollars)", omit.stat = c("f", "ser", "adj.rsq"),
          digits = 2, no.space = TRUE)
```

::: {.smaller-80 .muted}
Four cities, three coefficients. Where did Austin go? And what is the
constant?
:::

## Price by city: reading it {.smaller-90}

::::: {.columns}
::: {.column width="42%"}
Behind the scenes, R turns city into **dummy variables** (0/1), one per
city, and **drops one**: the **base level**, Austin (first alphabetically).

| Listing | City | Boston | LA | Miami |
|---|---|---|---|---|
| 1 | Austin | 0 | 0 | 0 |
| 2 | Boston | 1 | 0 | 0 |
| 3 | LA | 0 | 1 | 0 |
| 4 | Miami | 0 | 0 | 1 |
:::
::: {.column width="58%"}
- The **constant** (\$169.95) is the average price in the base city,
  **Austin**
- Each coefficient is the **difference from Austin**: Boston \$17.60
  cheaper, Los Angeles \$36.77 cheaper, Miami \$40.93 cheaper. All three
  precisely estimated (three stars): standard errors of \$2–3, because each
  city has thousands of listings

::: {.box-warn}
With small groups the same differences would not be significant. "Not
significant" can mean "no difference" or "not enough data": look at the
standard error and the group size before you conclude anything.
:::
:::
:::::

## Choosing the base level {.smaller}

```{r airbnb-relevel}
#| echo: true
airbnb$city <- relevel(factor(airbnb$city), ref = "Los Angeles")
m_city2 <- lm(price ~ city, data = airbnb)
stargazer(m_city, m_city2, type = "text", column.labels = c("base: Austin", "base: Los Angeles"),
          dep.var.labels = "Nightly price (dollars)", omit.stat = c("f", "ser", "adj.rsq"),
          digits = 2, no.space = TRUE)
```

::: {.smaller-80 .muted}
Same data, same fit (same $R^2$), different comparison. Pick the base your
reader cares about: the largest group, or the "status quo" option.
:::

# Regression Is Still Not Causation {.divider}

## What the coefficient does and does not say {.smaller-90}

Price is \$37 lower in Los Angeles than in Austin. Reviews go with slightly
lower prices. Facebook spend goes with more sales. All of these are
**associations**. None of them tells you what happens if you *change* $X$.

- A regression holds fixed **only the predictors you put in**. Everything
  you left out (listing quality, neighborhood, season) is still free to move
  with $X$ and $Y$ at the same time
- Adding predictors helps (newspaper's "effect" vanished once Facebook was
  in). It does not settle the question. Weeks 11–12 are about designs that
  do

::: {.box}
Words matter. Write "a one-unit increase in $X$ **is associated with** a
$\beta_1$ change in $Y$", not "$X$ **causes** $Y$", unless you have a reason
to believe it, and can say why.
:::

# Wrap-up {.divider}

## What's next

- **Tuesday, Sept. 22**: **logit**, regression for yes/no outcomes (will a
  listing get booked? will a customer churn?). Then the **Airbnb exercise**
  on your laptop: logs, more predictors, city and room type, and where you
  would buy a property. Bring the week 1 setup working
- **Homework 2** (due Thursday, Oct. 8): the same tools on a box-office
  dataset, OLS and logit, with `stargazer` tables and an executive memo

::: {.box}
Before Tuesday, run today's script on the Airbnb data and try one thing on
your own: replace `reviews_count` with `bedrooms` in `m1`. Predict the slope
first, then explain back what the intercept means.
:::

## Wrap-up

::: {.box}
Code and data to reproduce today's regressions, on the course website:

- [**`w4-code.zip`**](https://github.com/dadepro/mkt566/raw/main/w4/w4-code.zip):
  one-click download of everything below
- [**`w4-1-ols-class.R`**](https://github.com/dadepro/mkt566/blob/main/w4/code/w4-1-ols-class.R):
  every chart and regression shown today
- [**`data/airbnb.csv`**](https://github.com/dadepro/mkt566/blob/main/w4/code/data/airbnb.csv):
  the Airbnb dataset (also used in the Airbnb exercise on Tuesday, Sept. 22)
:::

::: {.smaller-80 .muted}
Required reading: Chapter
[3.4](https://bookdown.org/content/6ef13ea6-4e86-4566-b665-ebcd19d45029/secondarydata.html#linear-regression)
of *R for Marketing Students*. Optional: [Lecture 6 of Data Storytelling for
Marketers](https://raw.githack.com/dadepro/mkt-615/main/lectures/08-regression/08-regressions.html#1)
and Chapter 7 of *R for Marketing Research and Analytics*.
:::

# Questions? {.divider}
