---
title: "Regression for Yes/No Outcomes: Logit"
subtitle: "MKT 566 · Fall 2026"
author: "Davide Proserpio"
institute: "USC Marshall School of Business"
format:
  revealjs:
    theme: theme.scss
    slide-number: c/t
    width: 1280
    height: 720
    incremental: false
    transition: fade
    transition-speed: fast
    controls: true
    progress: true
    hash: true
    embed-resources: false
    link-external-newwindow: true
    footer: "MKT 566 · Marketing Analytics"
execute:
  echo: false
  warning: false
  message: false
fig-width: 7
fig-height: 4.2
fig-align: center
---

```{r setup}
#| include: false
library(ggplot2)
library(data.table)
library(scales)
library(stargazer)

options(width = 110)
set.seed(566)
knitr::opts_knit$set(root.dir = "code")

airbnb <- fread("code/data/airbnb.csv")
airbnb[, gem := as.integer(star_rating >= 4.5 & reviews_count > 20)]
lpm     <- lm(gem ~ price, data = airbnb)
lg1     <- glm(gem ~ price, data = airbnb, family = binomial)
m_logit <- glm(gem ~ price + guests_included + city + room_type, data = airbnb, family = binomial)
zero_at <- -coef(lpm)[1] / coef(lpm)[2]
```

# Before We Start {.divider}

## Housekeeping

- **Homework 2** (box office, OLS and logit) is out, due **Thursday, Oct. 8**.
  Everything it needs is in Thursday's deck and today's
- **Group project**: proposal presentations on **Oct. 13 and 15**
- Today: **logit** for about 40 minutes, then the **Airbnb case** on your
  laptop (aim to finish Task 5 in class; report due **Sunday, Sept. 27**)
- Thursday: **clustering**

## From "how much" to "whether" {.smaller-90}

Thursday's regressions predicted a **number**: sales, price. Many marketing
questions are **yes or no**:

- Will this customer **buy**? **Churn**? **Click** the ad? **Open** the email?
- Will this loan **default**? Will this listing be a **hit**?

::::: {.columns}
::: {.column width="50%"}
**Today you learn to**

1. See why a straight line is the wrong tool for a 0/1 outcome
2. Read a **logit** model: sign, odds, and predicted probabilities
3. Run it in R with `glm()` and put it in a `stargazer` table
:::
::: {.column width="50%"}
**What stays the same**

- Predictors, dummies, base levels, stars, "holding the others fixed": all
  as in OLS
- The AI writes the code. You choose the model and read the result
:::
:::::

## R script for today

In the `code/` folder on the course website, next to Thursday's:

1. **`w4-2-logit-class.R`**: every chart and model in today's deck
2. **`data/airbnb.csv`**: the same Airbnb listings

::: {.box}
[**`w4-code.zip`**](https://github.com/dadepro/mkt566/raw/main/w4/w4-code.zip)
has all of week 4: Thursday's two scripts, today's, and the data. If you
downloaded it on Thursday, you already have everything.
:::

# A Yes/No Outcome {.divider}

## Our outcome: is a listing a "gem"? {.smaller-90}

An investor wants listings that are **both** well rated **and** actually
booked. Define a **gem** as a listing with a rating of **4.5 or more** and
**more than 20 reviews**:

```{r gem-define}
#| echo: true
airbnb[, gem := as.integer(star_rating >= 4.5 & reviews_count > 20)]
table(airbnb$gem)
mean(airbnb$gem)
```

- `gem` is a **dummy**: 1 for a gem, 0 otherwise. About **20%** of listings
  are gems
- The question: **what makes a listing more likely to be a gem?** Start
  with price

::: {.smaller-80 .muted}
The `&` means both conditions must hold; `as.integer()` turns TRUE/FALSE
into 1/0, as in Task 7 of the case.
:::

## Share of gems by price

```{r gem-by-price}
#| fig-width: 8
#| fig-height: 4.2
airbnb[, pbin := cut(price, c(0, 50, 100, 150, 200, 300, 500, 1000, Inf),
                     labels = c("0-50", "50-100", "100-150", "150-200", "200-300",
                                "300-500", "500-1000", "1000+"))]
bins <- airbnb[, .(n = .N, share = mean(gem)), by = pbin][order(pbin)]
ggplot(bins, aes(x = pbin, y = share)) +
  geom_col(fill = "#990000", alpha = 0.85) +
  geom_text(aes(label = percent(share, accuracy = 1)), vjust = -0.4, size = 4) +
  scale_y_continuous(labels = percent, limits = c(0, 0.27)) +
  labs(title = "Share of Gems by Nightly Price", x = "Nightly price ($)", y = "Share of gems") +
  theme_minimal(base_size = 14)
```

::: {.smaller-80 .muted}
Same idea as the "average by group" charts of week 3, with a 0/1 variable:
the average of a dummy is a **share**. What do you see?
:::

## Share of gems by price: reading it {.smaller-90}

```{r gem-by-price-2}
#| fig-width: 8
#| fig-height: 3.3
ggplot(bins, aes(x = pbin, y = share)) +
  geom_col(fill = "#990000", alpha = 0.85) +
  scale_y_continuous(labels = percent) +
  labs(x = "Nightly price ($)", y = "Share of gems") +
  theme_minimal(base_size = 13)
```

- Gems are **rarer as price rises**: 22–23% of listings between $50 and
  $150, 12% between $500 and $1,000, 6% above $1,000. The very cheapest bin
  is also lower
- So the **probability** of being a gem depends on price. Can we model that
  probability with a regression, the way we modeled price on Thursday?

## Try the tool we have: a straight line {.smaller}

```{r lpm}
#| echo: true
#| eval: false
lpm <- lm(gem ~ price, data = airbnb)   # OLS, with a 0/1 outcome
```

::::: {.columns}
::: {.column width="58%"}
```{r lpm-plot}
#| fig-width: 6.5
#| fig-height: 4.6
sub <- airbnb[price <= 3000]
ggplot(sub, aes(x = price, y = gem)) +
  geom_jitter(height = 0.04, width = 0, alpha = 0.08, color = "darkgreen") +
  geom_abline(intercept = coef(lpm)[1], slope = coef(lpm)[2], color = "firebrick", linewidth = 1.1) +
  geom_hline(yintercept = c(0, 1), linetype = 3, color = "grey40") +
  scale_y_continuous(breaks = c(-0.1, 0, 0.5, 1), limits = c(-0.12, 1.06)) +
  labs(title = "Gem (0/1) vs. Price, with the OLS line", x = "Nightly price ($)", y = "Gem") +
  theme_minimal(base_size = 13)
```
:::
::: {.column width="42%"}
- The "linear probability model": the line is the **predicted
  probability** of being a gem
- Intercept **0.217**: a free listing has a 21.7% chance. Slope
  **−0.00009**: each extra dollar lowers it by 0.009 percentage points;
  \$100 more, by about **0.9 points**
- Reasonable so far. Now look at the right end of the line
:::
:::::

## Where the line breaks {.smaller-90}

```{r lpm-preds}
#| echo: true
pred <- predict(lpm)
sum(pred < 0)          # listings with a predicted probability below zero
round(min(pred), 2)    # the lowest one
```

- The line crosses zero at a price of about
  **$`r comma(round(zero_at))`** and keeps going: a $10,000 listing gets a
  "probability" of **−70%**
- A probability is trapped **between 0 and 1**. As it approaches either
  edge, the true relationship must **flatten**. A straight line cannot
  flatten: it falls at the same rate forever
- The slope has the same flaw: the line takes **0.9 points** off for every
  $100, whether a listing's chance of being a gem is at 20% (a cheap
  listing) or at 2% (a very expensive one). At 2% there are only 2 points
  left to lose, so near the edge the effect of price must shrink

::: {.box}
For a yes/no outcome we need a curve that is **linear in the middle** and
**flattens at both ends**. That curve is the logistic function, and the
model built on it is the **logit** (logistic regression).
:::

# Logistic Regression {.divider}

## The fix: squash the line into 0–1

::::: {.columns}
::: {.column width="45%"}
Keep the linear part, $z = \beta_0 + \beta_1 X$, and pass it through the
**logistic function**:

$$p = P(Y = 1 \mid X) = \frac{1}{1 + e^{-z}}$$

- Any $z$, from $-\infty$ to $+\infty$, comes out as a number between
  **0 and 1**
- $z = 0$ gives $p = 0.5$. Large positive $z$: close to 1. Large negative:
  close to 0
:::
::: {.column width="55%"}
```{r sigmoid}
#| fig-width: 6
#| fig-height: 4
zz <- data.table(z = seq(-6, 6, by = 0.05))[, p := 1 / (1 + exp(-z))]
ggplot(zz, aes(z, p)) +
  geom_hline(yintercept = c(0, 1), linetype = 3, color = "grey40") +
  geom_line(color = "#990000", linewidth = 1.3) +
  annotate("point", x = 0, y = 0.5, size = 3) +
  annotate("text", x = 0.4, y = 0.45, label = "z = 0, p = 0.5", hjust = 0, size = 4) +
  labs(title = "The logistic (S) curve", x = expression(z == beta[0] + beta[1]*X), y = "p") +
  theme_minimal(base_size = 14)
```
:::
:::::

## Odds and log-odds {.smaller-90}

The **odds** are the chance it happens, divided by the chance it does not.
Sports betting already speaks this language:

| Probability | Odds $= p / (1 - p)$ | In words |
|---|---|---|
| $p = 20\%$ (the chance a random listing is a gem) | $0.2 / 0.8 = 0.25$ | one to four: 1 gem for every 4 non-gems |
| $p = 50\%$ | $0.5 / 0.5 = 1$ | even: as likely as not |
| $p = 80\%$ | $0.8 / 0.2 = 4$ | four to one: 4 gems for every non-gem |

: {tbl-colwidths="[44,18,38]"}

- Rearranging the S-curve puts the model on this scale:
  $\dfrac{p}{1-p} = e^{z}$, and taking logs,
  $\log\left(\dfrac{p}{1-p}\right) = z = \beta_0 + \beta_1 X$
- The right-hand side is Thursday's **straight line**. That is what "logit"
  means, and why everything from OLS carries over: predictors, dummies,
  base levels, standard errors, stars

::: {.box}
A logit coefficient $\beta_1$ is the change in the **log-odds** for one more
unit of $X$. Nobody thinks in log-odds, so the next slide converts.
:::

## Reading a logit coefficient {.smaller-90}

The rule: one more unit of $X$ **multiplies the odds by** $e^{\beta_1}$.
R computes it for you (`exp(coef(m))`); your job is three steps:

| Step | Price example: $\beta_1 = -0.0008$ per dollar |
|---|---|
| **1. Sign and stars**, as in OLS | negative, significant: pricier listings are less likely to be gems |
| **2. Scale**: multiply $\beta_1$ by the change you care about | for \$100 more: $100 \times -0.0008 = -0.08$ |
| **3. Convert**: $e^{\beta_1}$ is what the odds are multiplied by | $e^{-0.08} = 0.92$: \$100 more multiplies the odds by 0.92, an **8% drop in the odds** |

- Shortcut, same as Thursday's log-level model: while the (scaled)
  coefficient stays within about $\pm 0.2$, just move the decimal:
  $-0.08 \to$ about $-8\%$. Bigger than that, use `exp()`

::: {.box-warn}
The percent change is in the **odds**, not in the probability. A 50%
increase in odds takes a probability of 20% to 27%, not to 30%: odds of
$0.25$ grow to $0.375$, and $p = 0.375 / 1.375 = 27\%$.
:::

## Logit on price: the curve, on the data

```{r logit-curve}
#| fig-width: 9
#| fig-height: 4.3
grid <- data.table(price = seq(0, 3000, by = 10))
grid[, `:=`(logit = predict(lg1, grid, type = "response"), line = predict(lpm, grid))]
mids <- data.table(price = c(25, 75, 125, 175, 250, 400, 750, 1500), share = bins$share, n = bins$n)
ggplot() +
  geom_hline(yintercept = 0, linetype = 3, color = "grey40") +
  geom_line(data = grid, aes(price, line, color = "OLS line"), linewidth = 1) +
  geom_line(data = grid, aes(price, logit, color = "Logit curve"), linewidth = 1.3) +
  geom_point(data = mids, aes(price, share, size = n), color = "darkgreen", alpha = 0.8) +
  scale_color_manual(values = c("OLS line" = "firebrick", "Logit curve" = "#0072B2"), name = NULL) +
  scale_size_area(max_size = 9, guide = "none") +
  scale_y_continuous(labels = percent) +
  labs(title = "Probability of Being a Gem vs. Price", x = "Nightly price ($)", y = "P(gem)") +
  theme_minimal(base_size = 14) + theme(legend.position = "top")
```

::: {.smaller-80 .muted}
Green dots: the observed share of gems in each price bin (sized by number
of listings). Which model fits the dots, and what happens past $2,000?
:::

## The curve: reading it {.smaller-90}

```{r logit-coef}
#| echo: true
lg1 <- glm(gem ~ price, data = airbnb, family = binomial)
round(coef(lg1), 5)
```

- The curve never dips below zero: at $2,372, where the line hits zero, it
  still says about **4%**; at $10,000, essentially zero. Where most
  listings are (\$50–\$300), curve and line agree almost perfectly
- The coefficient, with the recipe. **Sign and stars**: negative,
  significant: pricier listings are less likely to be gems. **Scale**, for
  \$100: $100 \times -0.00081 = -0.081$
- **Convert**: the shortcut says about an **8% drop in the odds** per
  \$100; the exact number, $e^{-0.081} = 0.92$, says **7.8%**. The
  shortcut was close enough

::: {.box}
Predicted probabilities are the reading a manager understands. `predict(...,
type = "response")` gives them: $50 → 21.5%, $200 → 19.6%, $500 → 16%,
$1,000 → 11%, $5,000 → 0.5%.
:::

## How it is estimated, and how to judge the fit {.smaller-90}

::::: {.columns}
::: {.column width="50%"}
**Estimation.** OLS picks the line with the smallest squared errors. Logit
picks the coefficients that make the observed 0s and 1s **most likely**
(**maximum likelihood**). R does it in a fraction of a second; you never
need the formula.

**Log-likelihood.** How likely the data is under the model. Always
negative; **closer to zero is better**. Use it to compare models on the
same data, the way you used R² on Thursday.
:::
::: {.column width="50%"}
**There is no R².** Alternatives:

- **Pseudo-R²** (McFadden): how much the log-likelihood improves over a
  model with no predictors. 0 to 1, but values of 0.1–0.3 are already decent
- **Accuracy**: predict "gem" when $p > 0.5$ and count the hits. Beware the
  base rate: saying "not a gem" for **everyone** is right **80%** of the
  time here. Week 9 (classifiers) covers this properly
:::
:::::

::: {.smaller-80 .muted}
For the price-only model: log-likelihood −25,650 vs. −25,697 without price;
pseudo-R² 0.002. Price matters, and explains very little, the same lesson
as Thursday.
:::

# Logit in R {.divider}

## One function: `glm()` {.smaller}

`glm()` works like `lm()`, with `family = binomial` to ask for a logit:

```{r glm-full-code}
#| echo: true
#| eval: false
m_logit <- glm(gem ~ price + guests_included + city + room_type, data = airbnb, family = binomial)
stargazer(m_logit, type = "text", omit.stat = c("f", "ser", "aic", "bic"),
          digits = 4, no.space = TRUE, single.row = TRUE)
```

::: {.stargazer-html}
```{r glm-full}
#| echo: false
#| results: asis
tab <- capture.output(
  stargazer(m_logit, type = "html", omit.stat = c("f", "ser", "aic", "bic"),
            digits = 4, no.space = TRUE, single.row = TRUE,
            dep.var.labels = "Gem (0/1)",
            covariate.labels = c("Price", "Guests included", "Boston",
                                 "Los Angeles", "Miami", "Private room",
                                 "Shared room")))
tab <- sub('<tr><td style="text-align:left"><em>Note:</em>.*$', '', tab)
tab <- sub("(Log Likelihood</td><td>-?[0-9,]+)\\.[0-9]+", "\\1", tab)
cat(tab, sep = "\n")
```
:::

::: {.smaller-80 .muted}
Same table, typeset for the slide; in your terminal it prints as text.
Standard errors in parentheses; stars: \*p < 0.1, \*\*p < 0.05,
\*\*\*p < 0.01. Read the signs, then the sizes. Which predictor has the
largest effect?
:::

## Reading the table {.smaller-90}

| Coefficient | Odds multiply by | Reading (holding the others fixed) |
|---|---|---|
| Price $-0.0017$ | $e^{-0.17} = 0.84$ per \$100 | \$100 more: odds of being a gem **16% lower** |
| Guests included $0.11$ | $1.12$ per guest | one more guest included: odds **12% higher** |
| Boston $0.17$, Los Angeles $0.16$ | $1.18$, $1.18$ | vs. Austin: odds about **18% higher** |
| Miami $-0.20$ | $0.82$ | vs. Austin: odds **18% lower** |
| Private room $-0.01$, not significant | $0.99$ | same odds as an entire home |
| Shared room $-0.71$ | $0.49$ | odds **halved** vs. an entire home |

- The **largest** effect is the shared-room dummy, not price: a \$100 price
  difference moves the odds less than the room type does
- Log-likelihood $-25{,}420$, vs. $-25{,}697$ with no predictors

## From odds to probabilities {.smaller-90}

Odds ratios compare. Probabilities decide. `predict()` with
`type = "response"` gives the probability for any listing you describe:

```{r predict-examples}
#| echo: true
examples <- data.frame(price = c(100, 100, 300, 100), guests_included = 2,
                       city = c("Austin", "Boston", "Austin", "Austin"),
                       room_type = c("Entire home", "Entire home", "Entire home", "Shared room"))
examples$p_gem <- round(predict(m_logit, examples, type = "response"), 3)
examples
```

- An entire home in Austin at \$100 for 2 guests: **21%** chance of being a
  gem. The same listing in Boston: **24%**. At \$300 instead of \$100:
  **16%**. As a shared room: **11%**
- This is the sentence for the investor: "moving the same property from
  Austin to Boston raises the chance it becomes a gem from 21% to 24%"

::: {.box}
Odds ratios for a table, probabilities for a decision. Always compute a few
predicted probabilities for realistic cases before you write a
recommendation.
:::

## OLS and logit, side by side {.smaller-90}

| | **OLS** (Thursday) | **Logit** (today) |
|---|---|---|
| Outcome | a number (price, sales) | yes/no (gem, buy, churn, click) |
| Model | $Y = \beta_0 + \beta_1 X$ | $\log\frac{p}{1-p} = \beta_0 + \beta_1 X$ |
| R | `lm(y ~ x)` | `glm(y ~ x, family = binomial)` |
| $\beta_1$ means | one more unit of $X$ → $\beta_1$ more units of $Y$ | one more unit of $X$ → odds $\times\, e^{\beta_1}$ |
| Predictions | `predict(m)` | `predict(m, type = "response")` gives probabilities |
| Fit | R², residual standard error | log-likelihood, pseudo-R² |
| Estimation | least squares | maximum likelihood |
| Same in both | predictors, logs of $X$, dummies and base levels, SE, stars, "holding the others fixed", **not causation** |

## Still not causation {.smaller-90}

Cheaper listings are more likely to be gems. Three stories, as always:

1. Lower prices bring more bookings, hence more reviews, hence gem status
2. Hosts of listings that are already popular have no need to discount, and
   hosts of struggling ones cut prices: the causality runs the other way
3. Gems are **older** listings (20+ reviews take time), and older listings
   were priced in an earlier, cheaper market

::: {.box}
"A \$100 higher price is **associated with** 16% lower odds of being a gem."
The model holds fixed guests, city, and room type. It does not hold fixed
listing age, photos, or host effort.
:::

# The Airbnb Case {.divider}

## The case {.smaller-90}

The handout
([**`w4-airbnb-case.html`**](https://raw.githack.com/dadepro/mkt566/main/w4/w4-airbnb-case.html))
works like RateBeer, with less help: an exact prompt for Task 1, skeletons
for Tasks 2–3, and only the goal from Task 4 on.

The client is an **investor** deciding where to buy and what to list:

1. **Tasks 1–3**: look at the data, price vs. reviews with and without logs,
   the three models side by side (Thursday, applied)
2. **Tasks 4–5**: more predictors, then city as a factor and as hand-made
   dummies
3. **Tasks 6–7**: a revenue proxy, room type, a quality dummy
4. **Task 8**: three recommendations, one of them a limitation

Under every task: **predict** before you run, **explain back** one line after.

::: {.box}
Aim to finish **Task 5** in class. Deliverable, same as always: an **R
Markdown report knitted to HTML** plus `ai-log.md`, due **Sunday, Sept. 27,
11:59 pm**. Data:
[**`airbnb-case-data.zip`**](https://github.com/dadepro/mkt566/raw/main/w4/airbnb-case-data.zip).
:::

## Wrap-up

::: {.box}
Code and data to reproduce today's models, on the course website:

- [**`w4-code.zip`**](https://github.com/dadepro/mkt566/raw/main/w4/w4-code.zip):
  both week 4 scripts and the data
- [**`w4-2-logit-class.R`**](https://github.com/dadepro/mkt566/blob/main/w4/code/w4-2-logit-class.R):
  every chart and model shown today
- [**The Airbnb case handout**](https://raw.githack.com/dadepro/mkt566/main/w4/w4-airbnb-case.html)
  and its [**data**](https://github.com/dadepro/mkt566/raw/main/w4/airbnb-case-data.zip)
:::

::: {.smaller-80 .muted}
Required reading: Chapter
[3.6](https://bookdown.org/content/6ef13ea6-4e86-4566-b665-ebcd19d45029/secondarydata.html#logistic)
of *R for Marketing Students*. Optional: Chapter 9.2 of *R for Marketing
Research and Analytics*.
:::

# Questions? {.divider}
