---
title: 'The Variation Case: Instructor Solution'
author: "MKT 566 · Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
---

<!-- INSTRUCTOR ONLY: keep out of w2/code/ and w2-code.zip. This file mirrors
     what a good student variation-report.Rmd should look like (one section per
     task, the chunk that makes the chart, a written answer), with model
     answers. It lives at w2/ root, so data is read from code/data/. A student
     file sits inside code/ and reads data/marketing_eda.csv instead. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
library(ggplot2)
library(dplyr)
library(scales)
df <- read.csv("code/data/marketing_eda.csv")
```

## Task 1: look at the data

```{r head}
head(df)
nrow(df)
```

**Answer:** There are 1,000 rows, one per customer: each row records who the
customer is (`Age`, `Gender`, `Device`), which ad channel reached them
(`Channel`), what was spent on them (`Ad_Spend`), and how they responded
(`Clicks`, `Purchases`, `Revenue`). `Gender`, `Device`, and `Channel` are
categorical; the rest are numeric (`CustomerID` is just an identifier).

## Task 2: summary statistics

```{r summary}
summary(df)
```

**Answer:** No missing values in any column. For `Ad_Spend`, the mean (216.8)
is well above the median (172.2). From the lecture: when the mean is pulled
above the median, the distribution is **right-skewed**; a minority of large
values drags the mean up while the median stays with the typical customer.

## Task 3: the distribution of ad spend

```{r adspend-hist}
ggplot(df, aes(Ad_Spend)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(aes(xintercept = mean(Ad_Spend)), linetype = "dashed") +
  geom_vline(aes(xintercept = median(Ad_Spend))) +
  labs(title = "Distribution of Ad Spend",
       subtitle = "Dashed = mean, solid = median",
       x = "Ad Spend ($)", y = "Count") +
  theme_minimal()
```

**Answer:** Clearly right-skewed: most customers receive modest spend, with a
long tail of customers receiving several times more. The dashed mean line sits
visibly to the right of the solid median line, which is exactly the skew the
summary statistics hinted at. A plausible marketing strategy behind this
shape: the firm spends a baseline amount on most customers and concentrates
extra budget on a small group it believes is high-value (retargeting,
lookalike audiences, big-ticket prospects).

## Task 4: the same chart, on a log scale

```{r adspend-hist-log}
ggplot(df, aes(Ad_Spend)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  scale_x_continuous(trans = "log10") +
  labs(title = "Distribution of Ad Spend (log scale)",
       x = "Ad Spend ($, log10)", y = "Count") +
  theme_minimal()
```

**Answer:** On the log scale the distribution becomes roughly symmetric and
bell-shaped, so the "typical" customer is easy to read off: about $150–200.
The low-spend customers, invisible in the left pile of the linear chart, are
now spread out and visible. This is the lecture's point about logs: they
expand small values, compress large ones, and turn a skewed distribution into
one you can actually read.

## Task 5: the channel mix

```{r channel-mix}
df %>%
  count(Channel) %>%
  mutate(p = n / sum(n)) %>%
  ggplot(aes(reorder(Channel, -n), n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent(p, accuracy = 0.1)), vjust = -0.3, size = 3.5) +
  labs(title = "Channel Mix", x = "Channel", y = "Customers") +
  theme_minimal()
```

**Answer:** Not balanced. Search dominates (34.7%), followed by Social
(22.7%), Display (19.4%), and Video (17.8%); Email is a distant last at 5.4%.
The mix is plausible for an online retailer: search captures customers with
existing intent, social and display do the prospecting, and email is cheap
but only reaches customers the firm already knows.

## Task 6: hunting outliers

```{r adspend-box}
ggplot(df, aes(y = Ad_Spend)) +
  geom_boxplot(fill = "lightblue") +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Boxplot of Ad Spend", y = "Ad Spend ($)") +
  theme_minimal()
```

```{r adspend-box-log}
ggplot(df, aes(y = log(Ad_Spend + 1))) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot of Ad Spend (log scale)", y = "log Ad Spend") +
  theme_minimal()
```

**Answer:** The 1.5 × IQR rule flags 39 customers, all on the high side; there
are no low outliers on the raw scale. On the log scale the box is nearly
symmetric and only a handful of extreme points remain flagged, telling us most
of the "outliers" are just the natural tail of a skewed distribution rather
than data errors. Should we drop them? Not without investigation: with 39
observations in 1,000, they can move averages. Best practice from the lecture:
run any later analysis with and without them, disclose whatever we do, and
prefer robust statistics (medians, quantiles) when describing spend.

## Task 7: my own question

*Example: is ad spend targeted by device?*

```{r spend-by-device}
ggplot(df, aes(Device, Ad_Spend, fill = Device)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(trans = "log10", labels = dollar_format()) +
  labs(title = "Ad Spend by Device (log scale)",
       x = NULL, y = "Ad Spend ($, log10)") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Answer:** I asked whether Mobile and Desktop customers receive different ad
spend. They essentially do not: the two distributions sit on top of each other
(means 214 vs. 221, medians 175 vs. 170). Spend does not appear to be targeted
by device in this data, even though mobile users are 65% of customers.

<!-- Common student variants for Task 7 and what they should find:
     - Age distribution: roughly uniform 18-65 (it was drawn that way), which
       is itself worth remarking on -- real customer bases are rarely uniform.
     - Distribution of Purchases: counts concentrated at 0-2 with a right
       tail; discrete, so a bar chart beats a histogram.
     - Clicks by Device: here there IS a difference (mobile clicks more),
       a nice contrast with spend being flat across devices. -->
