---
title: "Exploratory Data Analysis: Covariation"
subtitle: "MKT 566 · Fall 2026"
author: "Davide Proserpio"
institute: "USC Marshall School of Business"
format:
  revealjs:
    theme: theme.scss
    slide-number: c/t
    width: 1280
    height: 720
    incremental: false
    transition: fade
    transition-speed: fast
    controls: true
    progress: true
    hash: true
    embed-resources: false
    link-external-newwindow: true
    footer: "MKT 566 · Marketing Analytics"
execute:
  echo: false
  warning: false
  message: false
fig-width: 7
fig-height: 4.2
fig-align: center
---

```{r setup}
#| include: false
library(datarium)
library(ggplot2)
library(dplyr)
library(tidyr)
library(data.table)
library(patchwork)
library(scales)

options(width = 110)
set.seed(566)
knitr::opts_knit$set(root.dir = "code")

data("marketing", package = "datarium")
marketing <- as.data.table(marketing)
marketing[, total_ad_spend := youtube + facebook + newspaper]
df <- fread("code/data/marketing_eda.csv")
```

# Before We Start {.divider}

## Housekeeping

- **Homework 1** is due **Monday, Sept. 14**, on Brightspace. Same format
  as the case: an R Markdown report knitted to HTML, plus `ai-log.md`
- **Groups**: be in a group by **Tuesday, Sept. 15**. Add yours to the
  [sign-up sheet](https://docs.google.com/spreadsheets/d/1Jna-Noy_q3fSUGYgK7wW49jKpS5l3lKt2Q14QZ004xw/edit?usp=sharing)
- Today: the solution of the variation case, then **covariation**
- Thursday and next Tuesday: the **RateBeer case** (bring your laptop)

## "Am I actually learning anything?" {.smaller-90}

A common feeling in the first weeks of working this way: the AI writes code
you have never seen, you run it, and a chart appears. It feels like magic, not
learning. That feeling is **normal**, and it is not a sign that something is
wrong.

::::: {.columns}
::: {.column width="48%"}
**What this course does not teach**

- Writing R code from memory
- Knowing every R function

The AI does that, better than any of us.
:::
::: {.column width="52%"}
**What it does teach (and grade)**

1. **Choosing** the right chart or analysis for the question
2. **Asking** for it clearly
3. **Reading** the result: does it make sense? what does it say?
4. **Concluding**: what does it mean for the marketing decision?
:::
:::::

::: {.box}
Running code you could not write yourself is the point, not a problem. The
skill is to **read** it and **judge** the result. It takes a few weeks, and
the two habits on the next slide speed it up.
:::

## Two habits, starting this week {.smaller-90}

::::: {.columns}
::: {.column width="50%"}
**1. Predict before you run**

Before you make the chart, write one sentence:
*"I expect Search to have the most customers."*

Then look. Were you right? If not, is it the **data** or the **code**?

Without a guess, every chart looks "fine" and you learn nothing.
:::
::: {.column width="50%"}
**2. Explain it back**

After the chart works, pick the **one line** of code that does the work.
Say in your own words what it does.

Stuck? Ask the AI: *"Explain this line as if I have never coded."* Then say
it back in your words.

If you can explain it, you understand it.
:::
:::::

::: {.box}
Our next in-class exercise (the RateBeer case, Thursday) has a short box for
both under every task. This is the part of the report that shows **you** did
the thinking.
:::

## Reading ggplot: five pieces, every time {.smaller-90}

Every chart the AI writes has the same structure:

```r
ggplot(df, aes(x = Channel, y = Ad_Spend)) +   # 1. the data, 2. which column on which axis
  geom_boxplot() +                             # 3. the chart type
  facet_wrap(~ Device) +                       # 4. one panel per group (optional)
  labs(title = "Ad spend by channel",          # 5. labels
       x = NULL, y = "Ad spend ($)")
```

- `geom_point` = scatter, `geom_col`/`geom_bar` = bars, `geom_histogram`,
  `geom_boxplot`, `geom_line`, `geom_tile` = heatmap
- Everything after `labs()` is style (`theme_minimal()`, colors, scales)

::: {.box}
You do not need to **write** this. You need to **recognize** it. Then, when
the AI uses a `geom_` you did not ask for, or puts the wrong column on `x`,
you will notice.
:::

# The Variation Case: Debrief {.divider}

## Tasks 1–2: what is a row, and what do the summaries say?

```{r debrief-summary}
#| echo: true
df <- read.csv("data/marketing_eda.csv")
summary(df[, c("Age", "Ad_Spend", "Clicks", "Purchases", "Revenue")])
```

::: {.smaller-80 .muted}
One row = one customer. What did you conclude from this table?
:::

## Tasks 1–2: reading it

```{r debrief-summary-2}
summary(df[, c("Age", "Ad_Spend", "Clicks", "Purchases", "Revenue")])
```

::: {.smaller-90}
- **No missing values** (there would be an `NA's` row)
- `Ad_Spend`: mean **217**, median **172**. When the mean is above the
  median, there is a **right tail**: a few customers get much more spend
- `Purchases`: median **1**, max **7**. Most customers buy once or never
:::

## Task 3: the distribution of ad spend

```{r debrief-hist}
p_hist <- ggplot(df, aes(Ad_Spend)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(aes(xintercept = mean(Ad_Spend)), linetype = "dashed") +
  geom_vline(aes(xintercept = median(Ad_Spend))) +
  labs(title = "Distribution of Ad Spend", subtitle = "Dashed = mean, solid = median",
       x = "Ad spend ($)", y = "Customers") +
  theme_minimal()
p_hist
```

::: {.smaller-80 .muted}
What shape is this, and what marketing strategy could produce it?
:::

## Task 3: reading it

```{r debrief-hist-2}
p_hist
```

::: {.smaller-90}
- **Right-skewed**: the mean is to the right of the median, as the summary
  table suggested
- A likely strategy: a **basic** spend on everyone, plus extra budget on a
  few customers the firm thinks are high-value (retargeting, lookalike
  audiences)
:::

## Task 4: the same chart on a log scale

```{r debrief-hist-log}
p_hist_log <- ggplot(df, aes(Ad_Spend)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  scale_x_continuous(trans = "log10", labels = dollar) +
  labs(title = "Distribution of Ad Spend (log scale)", x = "Ad spend ($, log scale)",
       y = "Customers") +
  theme_minimal()
p_hist_log
```

::: {.smaller-80 .muted}
What can you see now that was invisible before?
:::

## Task 4: reading it

```{r debrief-hist-log-2}
p_hist_log
```

::: {.smaller-90}
- On the log scale the distribution is roughly **symmetric**. The "typical"
  customer is easy to see: **$150–200**
- The low-spend customers were all piled up on the left before. Now they are
  spread out. Logs stretch small values and squeeze large ones
:::

## Task 5: the channel mix

```{r debrief-channel}
p_channel <- df %>%
  count(Channel) %>%
  mutate(p = n / sum(n)) %>%
  ggplot(aes(reorder(Channel, -n), n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent(p, accuracy = 0.1)), vjust = -0.3, size = 4) +
  labs(title = "Channel Mix", x = NULL, y = "Customers") +
  theme_minimal()
p_channel
```

::: {.smaller-80 .muted}
Balanced? Does the mix make sense for an online retailer?
:::

## Task 5: reading it

```{r debrief-channel-2}
p_channel
```

::: {.smaller-90}
- **Not balanced**: Search is a third of customers, Email one in twenty
- It makes sense: search reaches customers who already **want to buy**,
  social and display **find new customers**, and email only reaches
  customers the firm already knows
:::

## Task 6: hunting outliers

```{r debrief-box}
#| fig-width: 10
#| fig-height: 4.2
p_box_raw <- ggplot(df, aes(x = "", y = Ad_Spend)) +
  geom_boxplot(fill = "lightblue") +
  scale_y_continuous(labels = dollar) +
  labs(title = "Ad Spend", x = NULL, y = "Ad spend ($)") +
  theme_minimal()
p_box_log <- ggplot(df, aes(x = "", y = Ad_Spend)) +
  geom_boxplot(fill = "lightblue") +
  scale_y_continuous(trans = "log10", labels = dollar) +
  labs(title = "Ad Spend (log scale)", x = NULL, y = "Ad spend ($, log scale)") +
  theme_minimal()
p_box_pair <- p_box_raw + p_box_log
p_box_pair
```

::: {.smaller-80 .muted}
How many outliers, on which side, and should we drop them?
:::

## Task 6: reading it

```{r debrief-box-2}
#| fig-width: 10
#| fig-height: 4.2
p_box_pair
```

::: {.smaller-90}
- The 1.5 × IQR rule flags **39** customers, all on the **high** side. On the
  log scale only a few remain. Most "outliers" are just the tail of a skewed
  distribution, not errors
- **Do not drop them** without a reason. Run later analyses **with and
  without** them, say what you did, and prefer medians to means
:::

## Task 7: your own question (two examples)

```{r debrief-task7}
#| fig-width: 10
#| fig-height: 4.2
p_t7a <- ggplot(df, aes(Device, Ad_Spend, fill = Device)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(trans = "log10", labels = dollar) +
  labs(title = "Ad Spend by Device (log scale)", x = NULL, y = "Ad spend ($)") +
  theme_minimal() + theme(legend.position = "none")
p_t7b <- ggplot(df, aes(Device, Clicks + 1, fill = Device)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(trans = "log10") +
  labs(title = "Clicks by Device (log scale)", x = NULL, y = "Clicks + 1") +
  theme_minimal() + theme(legend.position = "none")
p_t7 <- p_t7a + p_t7b
p_t7
```

::: {.smaller-80 .muted}
Same two groups, two different variables, both on a log scale. What is
different between the two panels?
:::

## Task 7: reading it {.nostretch}

```{r debrief-task7-2}
#| fig-width: 10
#| fig-height: 3.8
p_t7
```

::: {.smaller-90}
- **Spend** is the same on both devices; **clicks** are higher on mobile.
  The firm spends the same but gets more clicks on mobile
- Why "clicks + 1"? Six customers have **zero** clicks, and the log of zero
  does not exist. Adding 1 keeps them on the chart
:::

## What a strong answer looks like {.smaller-90}

**The question (Task 3):** *Is the distribution of ad spend symmetric or
skewed? What does the mean vs. the median tell you? What marketing strategy
might create this shape?*

::::: {.columns}
::: {.column width="48%"}
**Weak**

> "The distribution of ad spend is right-skewed. There are some outliers."

True, but it is just a caption. No numbers, no *why*, no *so what*.
:::
::: {.column width="52%"}
**Strong**

> "Right-skewed: the median customer gets **$172** but the mean is **$217**,
> so a few customers get several times the typical spend. The **39**
> outliers are all high-spend. That looks like a retargeting budget focused
> on a few customers, not a data error, so I would keep them and report
> medians."
:::
:::::

::: {.box}
The recipe: **the pattern → the number that shows it → the marketing
reason → what you would do.** Four steps, three sentences.
:::

::: {.smaller-80 .muted}
Full instructor solution:
[**`w2-1-variation-case-solution.html`**](https://raw.githack.com/dadepro/mkt566/main/w2/w2-1-variation-case-solution.html)
:::

## R script for today

In the `code/` folder on the course website:

1. **`w3-1-eda-covariation-class.R`**: reproduces every chart in the rest of
   today's deck, with a comment above each one
2. **`data/marketing_eda.csv`**: the week 2 case dataset, which we use again
   today

::: {.box}
Download everything as one zip:
[**`w3-code.zip`**](https://github.com/dadepro/mkt566/raw/main/w3/w3-code.zip),
open the folder in VS Code, and run along.
:::

::: {.smaller-80 .muted}
The debrief charts you just saw are in the variation case
[solution](https://raw.githack.com/dadepro/mkt566/main/w2/w2-1-variation-case-solution.html).
:::

# Covariation {.divider}

## The two questions of EDA

Last week's framing, one more time:

::: {.box}
1. What type of **variation** occurs *within* my variables? *(last week)*
2. What type of **covariation** occurs *between* my variables? *(today)*
:::

- Variation: one variable at a time (its distribution, its trend)
- Covariation: **two or more** variables, and how they change **together**

## What is covariation?

**Covariation** is when two or more variables change together in a related
way.

- The best way to see it is to **plot the relationship** between the
  variables
- Which plot to use depends, again, on the **type** of variables
- Most marketing questions are covariation questions: *does spend increase
  sales? do mobile users buy more? which channel converts best?*

## Which chart for which pair

| Pair of variables | Chart |
|---|---|
| **Continuous × continuous** (spend, sales) | Scatter plot (+ fitted line), correlation |
| **Categorical × continuous** (channel, purchases) | Bars of averages with error bars; boxplots by group |
| **Categorical × categorical** (channel, device) | Count table, heatmap |
| **Anything × time** | Line chart, one line per group |

: {tbl-colwidths="[50,50]"}

::: {.smaller-80 .muted}
Use this table when you write prompts: name the pair of variables, and you know what chart to ask for.
:::

# Two Continuous Variables {.divider}

## The `marketing` dataset again

Two hundred campaigns, ad spend on three channels, and sales:

```{r marketing-head}
#| echo: true
head(marketing)
```

::: {.smaller-80 .muted}
How would you visualize sales against ad spend?
:::

## Sales vs. total ad spend

```{r scatter}
#| echo: true
#| output-location: column
#| code-line-numbers: "1-2|4-5"
#| fig-width: 5.5
#| fig-height: 4.2
marketing$total_ad_spend <-
  marketing$youtube + marketing$facebook + marketing$newspaper
ggplot(marketing, aes(x = total_ad_spend, y = sales)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Sales vs. Total Ad Spend",
       x = "Total ad spend (thousands $)",
       y = "Sales (thousands of units)") +
  theme_minimal()
```

::: {.smaller-80 .muted}
Each point is one campaign. What do you see?
:::

## Sales vs. total ad spend: reading it

```{r scatter-2}
p_scatter <- ggplot(marketing, aes(x = total_ad_spend, y = sales)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "grey40") +
  labs(title = "Sales vs. Total Ad Spend", x = "Total ad spend (thousands $)",
       y = "Sales (thousands of units)") +
  theme_minimal()
p_scatter
```

::: {.smaller-90}
- A clear **positive** relationship: campaigns that spend more sell more
- Roughly a **straight line**, with a few high-spend campaigns that sold little
- The fitted line (`geom_smooth(method = "lm")`) is a preview of next week:
  **regression**
:::

## Can we do a more informative viz?

```{r scatter-facet}
#| echo: true
#| output-location: column
#| code-line-numbers: "1-4|9"
#| fig-width: 6
#| fig-height: 4.2
marketing_long <- pivot_longer(marketing,
  cols = c(youtube, facebook, newspaper),
  names_to = "channel",
  values_to = "ad_spend")
ggplot(marketing_long,
       aes(ad_spend, sales, color = channel)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~channel, scales = "free_x") +
  labs(title = "Sales vs. Spend by Channel",
       x = "Ad spend (thousands $)",
       y = "Sales") +
  theme_minimal() +
  theme(legend.position = "none")
```

::: {.smaller-80 .muted}
Same 200 campaigns, one panel per channel. What changed?
:::

## One panel per channel: reading it

```{r scatter-facet-2}
#| fig-width: 10
#| fig-height: 4.2
p_facet <- ggplot(marketing_long, aes(x = ad_spend, y = sales, color = channel)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "grey40") +
  facet_wrap(~channel, scales = "free_x") +
  scale_color_manual(values = c(youtube = "#FF0000", facebook = "#1877F2",
                                newspaper = "grey20")) +
  labs(title = "Sales vs. Ad Spend by Channel", x = "Ad spend (thousands $)",
       y = "Sales (thousands of units)") +
  theme_minimal() + theme(legend.position = "none")
p_facet
```

::: {.smaller-90}
- **YouTube**: strong, tight relationship. **Facebook**: positive but noisier.
  **Newspaper**: almost flat
- The total-spend chart **hid** this difference. Splitting a chart by a
  categorical variable (`facet_wrap`) is the most useful move in EDA
:::

## Correlation: the relationship as one number {.smaller-90}

The **correlation coefficient** $r$ measures how close two variables are to
a straight-line relationship, from −1 (perfect negative) through 0 (none) to
+1 (perfect positive):

```{r cor}
#| echo: true
cor(marketing$total_ad_spend, marketing$sales)
cor(marketing$youtube, marketing$sales)
cor(marketing$facebook, marketing$sales)
cor(marketing$newspaper, marketing$sales)
```

- Same ranking as the panels: YouTube (0.78), Facebook (0.58), Newspaper (0.23)

::: {.box-warn}
One number hides a lot. Anscombe's quartet is the classic example: four
datasets with the same means, the same $r = 0.82$, and the same fitted line,
and four completely different pictures. **Plot first, then summarize.**
:::

## Overplotting: when you have too many points

```{r blob}
#| fig-width: 6.5
#| fig-height: 4.2
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point() +
  labs(title = "Price vs. Carat (53,940 diamonds)", x = "Carat", y = "Price ($)") +
  theme_minimal()
```

::: {.smaller-80 .muted}
ggplot2's `diamonds` dataset, 53,940 rows. What is the problem with this chart?
:::

## Overplotting: two fixes

```{r blob-fixed}
#| fig-width: 11
#| fig-height: 4.2
p_alpha <- ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(color = "firebrick", se = FALSE) +
  labs(title = "alpha = 0.05 + smooth curve", x = "Carat", y = "Price ($)") +
  theme_minimal()
p_bins <- ggplot(diamonds, aes(x = carat, y = price)) +
  geom_bin2d(bins = 60) +
  scale_fill_viridis_c(labels = comma) +
  labs(title = "geom_bin2d(): count per cell", x = "Carat", y = "Price ($)",
       fill = "Diamonds") +
  theme_minimal()
p_alpha + p_bins
```

::: {.smaller-90}
- **Transparency** (`alpha`) + a smooth curve shows where most points are and
  the curved shape of the relationship
- **2D bins** (`geom_bin2d`) count the points in each cell: a heatmap. Use
  either one when a scatter plot turns into a blob (RateBeer: 290,000 reviews)
:::

# One Categorical and One Continuous Variable {.divider}

## Back to the case data

```{r df-head}
#| echo: true
df <- read.csv("data/marketing_eda.csv")
head(df)
```

::: {.smaller-80 .muted}
Which chart would you use to explore purchases by gender?
:::

## Purchases by gender

```{r bar-gender}
#| echo: true
#| output-location: column
#| code-line-numbers: "1-4|5-6"
#| fig-width: 5
#| fig-height: 4.2
avg_gender <- df %>%
  group_by(Gender) %>%
  summarise(avg_purchases = mean(Purchases),
            se_purchases = sd(Purchases) / sqrt(n()))
ggplot(avg_gender, aes(x = Gender, y = avg_purchases, fill = Gender)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c(F = "#E07A5F", M = "#3D5A80")) +
  labs(title = "Average Purchases by Gender",
       x = NULL, y = "Average purchases") +
  theme_minimal() + theme(legend.position = "none")
```

::: {.smaller-80 .muted}
First summarize (one row per group), then chart the summary. What do you conclude?
:::

## Purchases by gender: reading it

```{r bar-gender-2}
#| fig-width: 5
#| fig-height: 3
p_bar <- ggplot(avg_gender, aes(x = Gender, y = avg_purchases, fill = Gender)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c(F = "#E07A5F", M = "#3D5A80")) +
  labs(title = "Average Purchases by Gender", x = NULL, y = "Average purchases") +
  theme_minimal() + theme(legend.position = "none")
p_bar
```

::: {.smaller-90}
- Men average **1.09** purchases, women **1.01**. Is the difference
  **real**, or would another 1,000 customers flip it?
- The bars cannot tell you. We need to know how **precise** each average is
:::

## Add error bars

```{r bar-gender-se}
#| echo: true
#| output-location: column
#| code-line-numbers: "2-5"
#| fig-width: 5
#| fig-height: 4.2
p_bar +
  geom_errorbar(
    aes(ymin = avg_purchases - 1.96 * se_purchases,
        ymax = avg_purchases + 1.96 * se_purchases),
    width = 0.15, linewidth = 0.7)
```

::: {.smaller-80 .muted}
The **standard error**, $SE = sd / \sqrt{n}$, tells us how precise the
average is. Mean ± 1.96 × SE is a **95% confidence interval**: the range where
the true average most likely is. Now what do you conclude?
:::

## Error bars: reading it

```{r bar-gender-se-2}
#| fig-width: 5
#| fig-height: 4.2
p_bar_se <- p_bar +
  geom_errorbar(aes(ymin = avg_purchases - 1.96 * se_purchases,
                    ymax = avg_purchases + 1.96 * se_purchases),
                width = 0.15, linewidth = 0.7)
p_bar_se
```

::: {.smaller-90}
- The two intervals **overlap** a lot: the data cannot tell men and women
  apart on purchases
- Rule of thumb: bars of averages **without** error bars make small
  differences look real. Always ask for error bars
:::

## Add a third variable: the ad channel

```{r bar-gender-channel}
#| fig-width: 10
#| fig-height: 4
avg_gender_channel <- df %>%
  group_by(Gender, Channel) %>%
  summarise(avg_purchases = mean(Purchases),
            se_purchases = sd(Purchases) / sqrt(n()), n = n(), .groups = "drop")
p_bar_channel <- ggplot(avg_gender_channel, aes(x = Gender, y = avg_purchases, fill = Gender)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = avg_purchases - 1.96 * se_purchases,
                    ymax = avg_purchases + 1.96 * se_purchases),
                width = 0.15, linewidth = 0.7) +
  scale_fill_manual(values = c(F = "#E07A5F", M = "#3D5A80")) +
  facet_wrap(~Channel, nrow = 1) +
  labs(title = "Average Purchases by Gender and Ad Channel", x = NULL, y = "Average purchases") +
  theme_minimal() + theme(legend.position = "none")
p_bar_channel
```

::: {.smaller-80 .muted}
`facet_wrap(~Channel)` again. What do you see now?
:::

## Gender × channel: reading it

```{r bar-gender-channel-2}
#| fig-width: 10
#| fig-height: 4
p_bar_channel
```

::: {.smaller-90}
- The gap changes by channel: men buy more via **Search** (1.31 vs. 1.09),
  women via **Video** (1.20 vs. 0.94)
- **Email** has the widest error bars: only 54 customers. Small group, wide
  interval. Always check how many people are behind a bar
:::

## Averages hide distributions

```{r box-channel}
p_box_channel <- ggplot(df, aes(x = reorder(Channel, Ad_Spend, median),
                                y = Ad_Spend, fill = Channel)) +
  geom_boxplot(alpha = 0.8) +
  scale_y_continuous(trans = "log10", labels = dollar) +
  labs(title = "Ad Spend by Channel (log scale)", x = NULL, y = "Ad spend ($)") +
  theme_minimal() + theme(legend.position = "none")
p_box_channel
```

::: {.smaller-80 .muted}
One boxplot per group, sorted by median. Do the channels differ, and how?
:::

## Boxplots by group: reading it

```{r box-channel-2}
p_box_channel
```

::: {.smaller-90}
- The **medians** are almost the same: the typical customer gets the same
  spend in every channel
- The differences are in the **tails**: Video and Social have more
  high-spend outliers. An average would say "Video spends more". The boxplot
  says "Video spends more **on a few customers**". A different conclusion
:::

# Two Categorical Variables {.divider}

## Channel and device: how many customers in each cell?

```{r heat-counts}
#| fig-width: 8
#| fig-height: 3.8
counts_cd <- df %>% count(Channel, Device)
p_heat_n <- ggplot(counts_cd, aes(x = Channel, y = Device, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), color = "white", size = 6) +
  scale_fill_gradient(low = "#9ecae1", high = "#08306b") +
  labs(title = "Customers by Channel and Device", x = NULL, y = NULL, fill = "Customers") +
  theme_minimal()
p_heat_n
```

::: {.smaller-80 .muted}
A **heatmap**: color = count, number on top. What do you see?
:::

## Counts heatmap: reading it

```{r heat-counts-2}
#| fig-width: 8
#| fig-height: 3.8
p_heat_n
```

::: {.smaller-90}
- **Mobile** is the majority in every channel (65% overall)
- **Video** is the most mobile-heavy (125 vs. 53): video ads show up where
  people watch video
- A heatmap is just a **table with color**. For two categorical variables,
  that is usually all you need
:::

## Add a third variable: average ad spend per cell

```{r heat-spend}
#| fig-width: 8
#| fig-height: 3.8
avg_cd <- df %>% group_by(Channel, Device) %>%
  summarise(avg_adspend = mean(Ad_Spend), n = n(), .groups = "drop")
p_heat_spend <- ggplot(avg_cd, aes(x = Channel, y = Device, fill = avg_adspend)) +
  geom_tile(color = "white") +
  geom_text(aes(label = dollar(round(avg_adspend))), color = "white", size = 6) +
  scale_fill_gradient(low = "#9ecae1", high = "#08306b", labels = dollar) +
  labs(title = "Average Ad Spend by Channel and Device", x = NULL, y = NULL,
       fill = "Avg. spend") +
  theme_minimal()
p_heat_spend
```

::: {.smaller-80 .muted}
Same grid, the color is now the average spend in the cell. Anything stand out?
:::

## Spend heatmap: reading it, and the EDA loop

```{r heat-spend-2}
#| fig-width: 8
#| fig-height: 3.8
p_heat_spend
```

::: {.smaller-90}
- **Video / Desktop** at $315 is far above every other cell. Is it a real
  pattern, or a few outliers in a cell with only **53** customers?
- To find out, go back and make a boxplot of that cell (it is in the class
  script). This is the EDA loop: **notice → ask → check → refine**
:::

# Correlation Is Not Causation {.divider}

## Does ad spend *cause* sales? {.smaller-90}

Sales and ad spend move together strongly ($r = 0.87$). Three stories fit
that fact:

1. **Spend causes sales**: the ads work
2. **Sales cause spend**: managers set budgets as a share of expected sales,
   so big markets get big budgets *and* big sales
3. **Something else causes both**: the holiday season raises spend and sales
   at the same time

::: {.box-warn}
EDA can show you that two variables **move together**. It cannot tell you
**why**. Weeks 11–12 are about telling these stories apart.
:::

Until then, choose your words carefully: "spend **is associated with**
sales", not "spend **drives** sales".

# The RateBeer Case {.divider}

## The case {.smaller-90}

Thursday you get **292,680 beer reviews** from RateBeer (a beer-review site, now closed),
2000–2012: beer, brewery, style, alcohol content, five ratings, and a date.

The client: a **new craft brewery** deciding **which styles to brew and how
many**.

Two things make this harder than last week:

- The data is **dirty**: ratings stored as text like `"14/20"`, alcohol
  content with `"-"` for missing, dates stored as numbers, ciders and sake
  mixed in with the beers. Finding the problems is your job; fixing them is
  the AI's
- The questions are about **covariation**: rating by style, rating vs.
  alcohol, number of styles vs. rating, and how all of this changes over
  time

## How the case works {.smaller-90}

The handout
([**`w3-ratebeer-case.html`**](https://raw.githack.com/dadepro/mkt566/main/w3/w3-ratebeer-case.html))
works like the variation case, with less help:

1. **Tasks 1–3** (look, find the problems, clean): an exact prompt, then
   prompts with blanks
2. **Tasks 4–8** (variation warm-up, four covariation questions, time): only
   the goal, you write the prompt
3. **Task 9**: three recommendations to the brewery, each backed by a number

Under every task: **predict** before you run, **explain back** one line after.

Two sessions: **Thursday** Tasks 1–6, **Tuesday, Sept. 15** Tasks 7–9 and
the debrief. Deliverable, same as last week: an **R Markdown report knitted
to HTML** plus `ai-log.md`, due **Sunday, Sept. 20**.

::: {.box}
Bring your laptop with the week 1 setup working. Download the data
([**`ratebeer-case-data.zip`**](https://github.com/dadepro/mkt566/raw/main/w3/ratebeer-case-data.zip),
5 MB) before class.
:::

## Wrap-up

::: {.box}
Code and data to reproduce today's charts, on the course website:

- [**`w3-code.zip`**](https://github.com/dadepro/mkt566/raw/main/w3/w3-code.zip):
  one-click download of everything below
- [**`w3-1-eda-covariation-class.R`**](https://github.com/dadepro/mkt566/blob/main/w3/code/w3-1-eda-covariation-class.R):
  every chart shown today
- [**`data/marketing_eda.csv`**](https://github.com/dadepro/mkt566/blob/main/w3/code/data/marketing_eda.csv):
  the case dataset from week 2
- [**The RateBeer case handout**](https://raw.githack.com/dadepro/mkt566/main/w3/w3-ratebeer-case.html)
  and its [**data**](https://github.com/dadepro/mkt566/raw/main/w3/ratebeer-case-data.zip)
:::

::: {.smaller-80 .muted}
Required reading: [Chapter 7 of R for Data
Science](https://r4ds.had.co.nz/exploratory-data-analysis.html) (the
covariation section). Optional: Chapters 3–5 of *R for Marketing Research and
Analytics*.
:::

# Questions? {.divider}
