---
title: "The RateBeer Case: Debrief and Part 2"
subtitle: "MKT 566 · Fall 2026"
author: "Davide Proserpio"
institute: "USC Marshall School of Business"
format:
  revealjs:
    theme: theme.scss
    slide-number: c/t
    width: 1280
    height: 720
    incremental: false
    transition: fade
    transition-speed: fast
    controls: true
    progress: true
    hash: true
    embed-resources: false
    link-external-newwindow: true
    footer: "MKT 566 · Marketing Analytics"
execute:
  echo: false
  warning: false
  message: false
fig-width: 7
fig-height: 4.2
fig-align: center
---

```{r setup}
#| include: false
library(data.table)
library(ggplot2)
library(scales)
options(width = 110)

beer <- fread("data/ratebeer-sampled.csv.gz")
raw <- copy(beer)
rating_cols <- c("review_overall", "review_aroma", "review_appearance",
                 "review_palate", "review_taste")
for (col in rating_cols) beer[, (col) := as.numeric(sub("/.*", "", get(col)))]
beer[, beer_ABV := as.numeric(ifelse(beer_ABV == "-", NA, beer_ABV))]
beer[, review_date := as.Date(as.POSIXct(review_time, origin = "1970-01-01", tz = "UTC"))]
n_raw <- nrow(beer)
beer <- beer[!grepl("Sak|Cider|Perry|Mead", beer_style)]
```

# Before We Start {.divider}

## Housekeeping

- **Today** is the deadline for being in a group: add yours to the
  [sign-up sheet](https://docs.google.com/spreadsheets/d/1Jna-Noy_q3fSUGYgK7wW49jKpS5l3lKt2Q14QZ004xw/edit?usp=sharing)
- The **RateBeer report** (`ratebeer-report.html` + `ai-log.md`) is due
  **Sunday, Sept. 20, 11:59 pm** on Brightspace
- Thursday: **regression** (OLS), the tool behind every fitted line you drew
  last week

## Today's plan

1. **Debrief of Tasks 1–6** (about 25 minutes): chart, "what do you see?",
   then the reading. Compare with your own answers as we go
2. **Tasks 7–9** on your laptop: the brewery table, time, and the
   recommendation
3. Last 10 minutes: the **report checklist**

::: {.box}
Keep your `ratebeer-case.R` open. If a chart of yours looks different from
mine, that is worth a question, not a worry.
:::

# Tasks 1–3: Look, Find the Problems, Clean {.divider}

## Task 1: what is a row?

```{r t1}
#| echo: true
head(raw[, .(beer_brewerId, beer_style, beer_ABV, review_overall, review_time)], 4)
nrow(raw)
uniqueN(raw$beer_beerId); uniqueN(raw$beer_brewerId); uniqueN(raw$review_profileName)
```

::: {.smaller-80 .muted}
292,680 rows. What does one row represent, and why does it matter?
:::

## Task 1: reading it

- One row is one **review**: one person rating one beer once
- Behind the 292,680 rows: **50,826** beers, **5,841** breweries,
  **13,030** reviewers
- So "reviews by style" counts a beer **once per review**. A style with a few
  famous beers can look bigger than a style with many quiet ones

::: {.box}
The **unit of observation** is the first thing to settle in any dataset.
Every count and every average depends on it. We change it on purpose in Task 7.
:::

## Task 2: find the problems

```{r t2}
#| echo: true
for (col in c("beer_ABV", "review_overall", "review_time")) {
  cat("\n==", col, "(", class(raw[[col]]), ")\n")
  print(head(sort(table(raw[[col]]), decreasing = TRUE), 5))
}
```

::: {.smaller-80 .muted}
Three columns, three problems. What are they?
:::

## Task 2: reading it {.smaller-90}

- **Ratings are text**: `"14/20"` is five characters, not a number. R cannot
  average text
- **`beer_ABV` has `"-"`** in 14,140 rows. It means **missing**, not zero.
  Real alcohol-free beers are in the data at 0.0–0.5, so 14,140 zeros would
  be wrong and would pull the average down
- **`review_time` is a Unix timestamp**: seconds since January 1, 1970.
  1157587200 is September 7, 2006
- And in `beer_style`: sake (12 kinds), cider, perry, and mead are not beer;
  `Klsch` and `&#40;IPA&#41;` are encoding damage from the scraper

::: {.box}
None of this is visible in a chart. You find it by **looking at the values**
before you plot anything. That is what Tasks 1–2 are for.
:::

## Task 3: after cleaning

```{r t3}
#| echo: true
n_raw - nrow(beer); nrow(beer)
summary(beer[, .(review_overall, review_taste, review_appearance, beer_ABV)])
range(beer$review_date)
```

::: {.smaller-80 .muted}
What is the scale of each rating? How many ABV values are missing?
:::

## Task 3: reading it {.smaller-90}

- Non-beers removed: **3,693** rows, leaving **288,987**
- Scales differ: overall **1–20**, taste and aroma **1–10**, appearance and
  palate **1–5**. A taste of 7 and an overall of 14 are both 70% of their
  scale; you cannot compare the raw numbers
- **14,019** reviews have no ABV: keep them for everything about ratings,
  drop them only in ABV charts (`na.rm = TRUE`)
- Reviews run from April 2000 to **January 13, 2012**. Remember that date
  for Task 8

::: {.box-warn}
Did you check the AI's work? Mean overall should be **13.2**. If yours was
0.7, the assistant kept the `/20` part. The assistant cannot see your output.
:::

# Task 4: Variation Warm-up {.divider}

## Reviews by style

```{r t4-styles}
#| fig-height: 4.6
style_n <- beer[, .N, by = beer_style][order(-N)][1:20]
style_n[, beer_style := gsub("&#40;", "(", gsub("&#41;", ")", beer_style))]
p_styles <- ggplot(style_n, aes(x = reorder(beer_style, N), y = N)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Reviews by Style (top 20 of 85)", x = NULL, y = "Reviews") +
  theme_minimal()
p_styles
```

::: {.smaller-80 .muted}
Which styles dominate? And what does "most reviewed" actually measure?
:::

## Reviews by style: reading it

```{r t4-styles-2}
#| fig-height: 4.2
p_styles
```

::: {.smaller-90}
- **IPA** leads (18,635), then Pale Lager, Belgian Strong Ale, Imperial Stout
- Pale Lager is the most **consumed** beer in the world, and it is second
  here only because everyone has tried one. This is what **enthusiasts**
  write about, not what the market buys. Keep that in mind for Task 9
:::

## The distribution of ABV

```{r t4-abv}
p_abv <- ggplot(beer[!is.na(beer_ABV)], aes(beer_ABV)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "grey20") +
  labs(title = "Distribution of ABV", x = "ABV (%)", y = "Reviews") +
  theme_minimal()
p_abv
```

::: {.smaller-80 .muted}
What shape? And what is hiding at the far right?
:::

## The distribution of ABV: reading it

```{r t4-abv-2}
p_abv
```

::: {.smaller-90}
- **Right-skewed**: most reviews are of beers between 4% and 8%, with a long
  tail past 10%
- The maximum is **57.7%**: Schorschbräu Schorschbock, a real freeze-distilled
  Eisbock, not a typo. An outlier is not automatically an error. **Check
  before you drop**
:::

# Task 5: Rating by Style {.divider}

## Average rating by style

```{r t5}
#| fig-height: 5.2
style_stats <- beer[, .(avg = mean(review_overall), se = sd(review_overall) / sqrt(.N),
                        n = .N, abv = mean(beer_ABV, na.rm = TRUE)), by = beer_style][n >= 500]
style_stats[, beer_style := gsub("&#40;", "(", gsub("&#41;", ")", beer_style))]
style_ends <- style_stats[order(-avg)][c(1:10, (.N - 9):.N)]
style_ends[, group := rep(c("Top 10", "Bottom 10"), each = 10)]
p_t5 <- ggplot(style_ends, aes(x = reorder(beer_style, avg), y = avg, fill = group)) +
  geom_col() +
  geom_errorbar(aes(ymin = avg - 1.96 * se, ymax = avg + 1.96 * se), width = 0.3) +
  coord_flip() +
  scale_fill_manual(values = c("Top 10" = "steelblue", "Bottom 10" = "grey60")) +
  labs(title = "Rating by Style: Top and Bottom 10 (styles with 500+ reviews)",
       x = NULL, y = "Average overall rating (out of 20)", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
p_t5
```

::: {.smaller-80 .muted}
Top three, bottom three? Why are the error bars so small? (Your chart has all 50 styles; this shows the two ends.)
:::

## Average rating by style: reading it {.smaller-90}

```{r t5-table}
style_stats[order(-avg)][c(1:3, (.N - 2):.N), .(beer_style, avg = round(avg, 2), n, abv = round(abv, 1))]
```

- **Imperial Stout, Quadrupel, Double IPA** at the top (15.4–15.9); **Pale
  Lager, Malt Liquor, Low Alcohol** at the bottom (6.4–7.4). Less than half
- Thousands of reviews per style, so the error bars are tiny: these gaps are
  not noise. The 500-review rule keeps a style with 12 lucky reviews off the
  top
- Look at the ABV column: the top styles are all **strong** (9–11%), the
  bottom ones light. Style and alcohol are tangled together. Next task

# Task 6: Rating vs. Alcohol {.divider}

## First attempt: a scatter plot

```{r t6-blob}
ggplot(beer[!is.na(beer_ABV)], aes(beer_ABV, review_overall)) +
  geom_point(alpha = 0.02, size = 0.5) +
  labs(title = "Overall Rating vs. ABV (275,000 reviews)", x = "ABV (%)", y = "Overall rating") +
  theme_minimal()
```

::: {.smaller-80 .muted}
Even with transparency. What is the problem, and what did you ask for instead?
:::

## Second attempt: average by ABV

```{r t6-binned}
abv_bins <- beer[!is.na(beer_ABV) & beer_ABV <= 15,
                 .(avg = mean(review_overall), n = .N), by = .(abv = round(beer_ABV))]
p_t6 <- ggplot(abv_bins, aes(abv, avg)) +
  geom_line(color = "grey50") +
  geom_point(aes(size = n), color = "steelblue") +
  scale_size_continuous(labels = comma) +
  labs(title = "Average Overall Rating by ABV", x = "ABV (%, rounded)",
       y = "Average overall rating", size = "Reviews") +
  theme_minimal()
p_t6
```

::: {.smaller-80 .muted}
Correlation: 0.40. Describe the relationship. Does stronger **cause** better?
:::

## Rating vs. alcohol: reading it

```{r t6-binned-2}
p_t6
```

::: {.smaller-90}
- Ratings **rise** from about 11/20 at 3–4% to 15/20 at 9–10%, then
  **flatten**. Past 13% the points bounce: few beers live there (small dots)
- Overplotting fix from Tuesday: 275,000 points became 16 averages
- Causal? A 12% imperial stout is brewed with more care and reviewed by people
  who **sought it out**. "Stronger beers are rated higher" is a fact.
  "Alcohol makes beer better" is a story
:::

## What a strong answer looks like {.smaller-90}

**The question (Task 6):** *Describe the relationship. Does it flatten?
Does higher alcohol cause higher ratings?*

::::: {.columns}
::: {.column width="46%"}
**Weak**

> "There is a positive relationship between ABV and rating. Correlation is
> not causation."

Both sentences are true. Neither says anything about *this* data.
:::
::: {.column width="54%"}
**Strong**

> "Ratings climb from **11/20** at 4% to **15/20** at 10% ($r$ = **0.40**),
> then flatten; above 13% there are too few beers to say. I would not call
> this causal: strong beers are mostly imperial stouts and quadrupels, made
> with more ingredients and bought on purpose by enthusiasts, so style and
> reviewer both change with ABV."
:::
:::::

::: {.box}
The recipe again: **the pattern → the number → the reason → what it means for
the decision.**
:::

# Part 2: Tasks 7–9 {.divider}

## What you do now {.smaller-90}

::::: {.columns}
::: {.column width="50%"}
**Task 7: the brewery table**

One row per **brewery**: number of styles, number of reviews, average rating.
Keep breweries with 10+ reviews. Styles vs. rating, with a fitted line.

*Watch for:* the unit of observation just changed. Where did the 289,000
reviews go?

**Task 8: over time**

Reviews per year and average rating per year.

*Watch for:* the data ends on **January 13, 2012**.
:::
::: {.column width="50%"}
**Task 9: the recommendation**

Three bullets for a new craft brewery: which styles, how many. Each bullet
rests on **one number** from a chart. One bullet must state a **limitation**
of this data.

*Hint:* who writes reviews on RateBeer, and what does the data **not**
contain (prices, volumes, costs)?

*Optional extra chart:* the share of reviews going to IPAs by year.
:::
:::::

::: {.box}
Predict first, then run, then explain one line back. Write the prediction
**before** the chart; a wrong prediction is fine and gets full credit.
:::

## Give your assistant a memory {.smaller-90}

Every tool reads one file at the start of each chat: `CLAUDE.md` (Claude
Code), `AGENTS.md` (Codex), `.github/copilot-instructions.md` (Copilot),
`.cursor/rules` (Cursor). Standing instructions go there **once**.

The handout's Setup step 4 has the prompt. It tells the assistant, among
other things:

- I am a beginner: comment each step, do not run code, I run it
- Print every chart in the chunk, never only `ggsave()`
- **After every request, append one line to `ai-log.md`**

::: {.box}
That last rule makes the AI log write itself. No more "remember before you
close the chat". Two minutes now, and it works for every assignment this
semester (copy the file into each new project folder).
:::

::: {.smaller-80 .muted}
Copilot follows the file only in agent mode. Copilot chat users keep the
end-of-session log prompt.
:::

## Before you submit: the checklist {.smaller-90}

Open your knitted `ratebeer-report.html` in a browser and check:

1. **Name and 10-digit USC ID** under the title
2. **Every chart is visible inside the HTML.** Code but no chart means the
   chunk only saved a file. Ask the assistant to print the chart in the chunk.
   The grader receives one file, not your figures folder
3. Under every task: **prediction**, **answer** with a number, **explain it
   back** in your own words
4. `ai-log.md` written by the assistant **before** you closed the chat

**Submit** both files on Brightspace by **Sunday, Sept. 20, 11:59 pm**.

::: {.smaller-80 .muted}
Handout:
[**`w3-ratebeer-case.html`**](https://raw.githack.com/dadepro/mkt566/main/w3/w3-ratebeer-case.html).
Thursday: regression.
:::

# Questions? {.divider}
