---
title: 'The RateBeer Case: Instructor Solution'
author: "MKT 566 · Fall 2026"
output:
  html_document:
    toc: true
    toc_float: true
---

<!-- INSTRUCTOR ONLY: keep out of w3/code/ and the zips. Mirrors what a good
     student ratebeer-report.Rmd should look like (one section per task, the
     code chunk, prediction / answer / explain-it-back), with model answers.
     Reads the gz dataset from data/, so it knits from w3/ root. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
library(data.table)
library(ggplot2)
library(scales)
beer <- fread("data/ratebeer-sampled.csv.gz")
```

## Task 1: look at the data

```{r look}
head(beer)
nrow(beer); ncol(beer)
str(beer)
uniqueN(beer$beer_beerId); uniqueN(beer$beer_brewerId); uniqueN(beer$review_profileName)
```

**My prediction:** one row is one review.

**My answer:** 292,680 rows and 12 columns. One row is one **review** of one
beer by one reviewer, not one beer: there are 50,826 distinct beers, 5,841
breweries, and 13,030 reviewers. So "reviews by style" counts a beer once per
review; a style with a few heavily reviewed beers can look bigger than a style
with many lightly reviewed ones.

**Explain it back:** `fread()` (from data.table) reads a CSV like `read.csv()`
but is many times faster on large files, and returns a data.table.

## Task 2: find the dirt

```{r dirt}
for (col in c("beer_ABV", "review_overall", "review_time", "beer_style")) {
  cat("\n==", col, "(", class(beer[[col]]), ")\n")
  print(head(sort(table(beer[[col]]), decreasing = TRUE), 6))
}
sort(unique(beer$beer_style))
```

**My answer:**

- The five `review_` columns are **text** like `"14/20"`: the score and its
  scale glued together. R cannot average text.
- `beer_ABV` is text too, because 14,140 rows contain `"-"`. It means
  **missing**, not zero: real non-alcoholic beers appear in the data with
  ABV 0.0–0.5, and treating 14,140 missing values as zero would pull the
  mean ABV from 6.6 down to about 6.3 and put 14,000 beers in the
  "alcohol-free" bin.
- `review_time` is a **Unix timestamp**: seconds since January 1, 1970.
  1157587200 is September 7, 2006.
- `beer_style` includes sake (12 styles starting with `Sak`), cider, perry,
  and mead. Several names are garbled by encoding: `Klsch` (Kölsch),
  `Oktoberfest/Mrzen` (Märzen), `Bire de Garde`, and HTML entities such as
  `&#40;IPA&#41;` for "(IPA)". Cosmetic, but a reminder that the data went
  through a scraper.

**Explain it back:** a number column stores quantities R can add and average;
a text column stores characters. `"14/20"` is a string of five characters, so
`mean()` has nothing to compute until we extract the 14.

## Task 3: clean it

```{r clean}
rating_cols <- c("review_overall", "review_aroma", "review_appearance",
                 "review_palate", "review_taste")
for (col in rating_cols) beer[, (col) := as.numeric(sub("/.*", "", get(col)))]
beer[, beer_ABV := as.numeric(ifelse(beer_ABV == "-", NA, beer_ABV))]
beer[, review_date := as.Date(as.POSIXct(review_time, origin = "1970-01-01", tz = "UTC"))]
n_before <- nrow(beer)
beer <- beer[!grepl("Sak|Cider|Perry|Mead", beer_style)]
n_before - nrow(beer); nrow(beer)
summary(beer[, c(rating_cols, "beer_ABV", "review_date"), with = FALSE])
sum(is.na(beer$beer_ABV))
```

**My answer:** overall is on a 1–20 scale, taste and aroma 1–10, appearance
and palate 1–5, so a taste of 7 is 70% of its scale while an overall of 14 is
70% of a different one; they can only be compared after rescaling. Removing
non-beers dropped 3,693 rows, leaving 288,987. 14,019 reviews have a missing
ABV: keep them for everything about ratings, and drop them only in the ABV
charts (that is what `na.rm = TRUE` does). Reviews run from April 2000 to
January 13, 2012.

**Explain it back:** `sub("/.*", "", x)` deletes the slash and everything after
it, turning `"14/20"` into `"14"`; `as.numeric()` then makes it a number.

## Task 4: variation warm-up

```{r style-bars}
style_n <- beer[, .N, by = beer_style][order(-N)][1:20]
ggplot(style_n, aes(x = reorder(beer_style, N), y = N)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Reviews by Style (top 20)", x = NULL, y = "Reviews") +
  theme_minimal()
```

```{r abv-hist}
ggplot(beer[!is.na(beer_ABV)], aes(beer_ABV)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "grey20") +
  labs(title = "Distribution of ABV", x = "ABV (%)", y = "Reviews") +
  theme_minimal()
beer[which.max(beer_ABV), .(beer_name, beer_style, beer_ABV)]
```

**My prediction:** IPA most reviewed; typical ABV around 5–6%.

**My answer:** IPAs lead (18,635 reviews), then Pale Lager, Belgian Strong
Ale, and Imperial Stout. "Most reviewed" measures what RateBeer's users, who
are enthusiasts, like to write about: Pale Lager is by far the most
*consumed* style in the world and it is second here only because everybody
has tried one. For a craft brewery, review volume is a proxy for enthusiast
attention, not sales. ABV is right-skewed: most reviews are of beers between
4% and 8%, with a long tail. The maximum, Schorschbräu Schorschbock at
57.7%, is a real product (a freeze-distilled Eisbock that held the world
record), not a typo. Outliers are not necessarily errors; check before
dropping.

**Explain it back:** `reorder(beer_style, N)` sorts the style labels by their
review count, so the bars come out in order instead of alphabetically.

## Task 5: rating by style

```{r rating-by-style, fig.height=6}
style_stats <- beer[, .(avg = mean(review_overall), se = sd(review_overall) / sqrt(.N),
                        n = .N, abv = mean(beer_ABV, na.rm = TRUE)), by = beer_style][n >= 500]
ggplot(style_stats, aes(x = reorder(beer_style, avg), y = avg)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = avg - 1.96 * se, ymax = avg + 1.96 * se), width = 0.3) +
  coord_flip() +
  labs(title = "Average Overall Rating by Style (styles with 500+ reviews)",
       x = NULL, y = "Average overall rating (out of 20)") +
  theme_minimal()
style_stats[order(-avg)][c(1:3, (.N - 2):.N), .(beer_style, avg = round(avg, 2), n, abv = round(abv, 1))]
```

**My prediction:** IPAs highest, light lagers lowest.

**My answer:** Imperial Stout (15.9), Abt/Quadrupel (15.7), and Imperial/Double
IPA (15.4) top the list; Pale Lager (7.4), Malt Liquor (6.8), and Low Alcohol
(6.4) are at the bottom, less than half the top scores. The minimum-review
rule keeps a style with 12 lucky reviews from topping the chart; with
thousands of reviews per style the error bars are tiny, so these differences
are not noise. The top styles are all strong (9–11% ABV), dark or intense;
the bottom styles are light lagers around 5% or weaker. Style and alcohol
content are confounded, which is exactly the next task.

**Explain it back:** the standard error, `sd / sqrt(n)`, says how much the
average would move if we drew another sample of reviews; mean ± 1.96 × SE is
the range (95% confidence interval) where the true average plausibly sits.

## Task 6: rating vs. alcohol

```{r abv-scatter, fig.height=4}
ggplot(beer[!is.na(beer_ABV)], aes(beer_ABV, review_overall)) +
  geom_point(alpha = 0.02, size = 0.5) +
  labs(title = "Overall Rating vs. ABV (all reviews)", x = "ABV (%)", y = "Overall rating") +
  theme_minimal()
```

```{r abv-binned, fig.height=4}
abv_bins <- beer[!is.na(beer_ABV) & beer_ABV <= 15,
                 .(avg = mean(review_overall), n = .N), by = .(abv = round(beer_ABV))]
ggplot(abv_bins, aes(abv, avg)) +
  geom_line(color = "grey50") +
  geom_point(aes(size = n), color = "steelblue") +
  scale_size_continuous(labels = comma) +
  labs(title = "Average Overall Rating by ABV", x = "ABV (%, rounded)",
       y = "Average overall rating", size = "Reviews") +
  theme_minimal()
cor(beer$beer_ABV, beer$review_overall, use = "complete.obs")
```

**My prediction:** stronger beers rated higher.

**My answer:** The raw scatter is a blob: 275,000 points on a grid of
integer ratings, so nothing is visible even with transparency. Averaging
within each ABV value shows a strong positive relationship: from about 11/20
at 3–4% to 15/20 at 9–10%, with a correlation of 0.40. Above 10% the curve
flattens; beyond 13% the averages bounce around because few beers exist
there. Causal? Not necessarily: strong beers are mostly imperial stouts,
quadrupels, and barley wines, brewed with more ingredients and care, and
reviewed by enthusiasts who sought them out. A 12% beer is rarely a casual
purchase, so its reviews come from people predisposed to like it.

**Explain it back:** `by = .(abv = round(beer_ABV))` groups every review by
its ABV rounded to the nearest percent and averages the ratings inside each
group, turning 275,000 points into 16, which is why the chart became readable.

## Task 7: the brewery table

```{r brewery}
brewer <- beer[, .(n_styles = uniqueN(beer_style), n_reviews = .N,
                   avg = mean(review_overall)), by = beer_brewerId][n_reviews >= 10]
nrow(brewer); summary(brewer$n_styles)
ggplot(brewer, aes(n_styles, avg)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "firebrick") +
  labs(title = "Breweries: Number of Styles vs. Average Rating (10+ reviews)",
       x = "Distinct styles", y = "Average overall rating") +
  theme_minimal()
cor(brewer$n_styles, brewer$avg)
brewer[order(-avg)][1]; brewer[order(avg)][1]
```

**My prediction:** more styles, higher rating.

**My answer:** 2,439 breweries with at least 10 reviews; the median brewery
has 8 styles, a quarter have 13 or more. Breadth and rating are positively
related (correlation 0.35): breweries with 20+ styles average about 13.7
versus 10.6 for single-style breweries. Two stories other than "variety makes
you good": survivorship (only breweries that were already good and popular
lasted long enough to launch many styles), and reverse causality (good
ratings bring demand, and demand funds new styles). Highest-rated brewery
with 10+ reviews: id 623 (17.4 on 422 reviews); lowest: id 4603 (5.5 on 48).

**Explain it back:** `by = beer_brewerId` collapses the 289,000 review rows
into one row per brewery, computing the three summaries inside each group;
the individual reviews are gone from this table, only their summaries remain.

## Task 8: over time

```{r time, fig.height=3.8}
by_year <- beer[, .(reviews = .N, avg = mean(review_overall)), by = .(year = year(review_date))][order(year)]
ggplot(by_year, aes(year, reviews)) + geom_line() + geom_point() +
  scale_y_continuous(labels = comma) + scale_x_continuous(breaks = 2000:2012) +
  labs(title = "Reviews per Year", x = NULL, y = "Reviews") + theme_minimal()
ggplot(by_year, aes(year, avg)) + geom_line() + geom_point() +
  scale_x_continuous(breaks = 2000:2012) +
  labs(title = "Average Overall Rating per Year", x = NULL, y = "Average overall rating") +
  theme_minimal()
range(beer$review_date)
```

**My prediction:** ratings flat.

**My answer:** Reviews grow from 461 in 2000 to 45,112 in 2011, then
"collapse" to 1,355 in 2012, because the data ends on January 13, 2012: the
last year has 13 days, not a crisis. Reporting a partial period as a drop is
the most common trend-chart mistake. The average rating drifts up from 12.3
(2001) to 13.5 (2011). Two non-beer explanations: the reviewer population
changed (early users were hardcore critics, later ones more generous), and
the mix of beers changed (the craft boom means more imperial stouts and IPAs,
which score higher, in the later years).

**Explain it back:** `year(review_date)` extracts the four-digit year from a
date, so `by = year` groups all reviews posted in the same calendar year.

## Task 9: the recommendation

```{r ipa-share, fig.height=3.8}
share <- beer[year(review_date) %in% 2002:2011,
              .(share = mean(beer_style %in% c("India Pale Ale &#40;IPA&#41;", "Imperial/Double IPA"))),
              by = .(year = year(review_date))][order(year)]
ggplot(share, aes(year, share)) + geom_line() + geom_point() +
  scale_y_continuous(labels = percent) + scale_x_continuous(breaks = 2002:2011) +
  labs(title = "Share of Reviews Going to IPAs and Double IPAs", x = NULL, y = "Share of reviews") +
  theme_minimal()
```

1. **Lead with a strong, intense style.** Imperial stouts, quadrupels, and
   double IPAs average 15.4–15.9/20 versus 7.4 for pale lager (Task 5), and
   ratings rise with ABV up to about 10% (Task 6). A flagship in the 8–10%
   range plays where enthusiasts are most generous.
2. **Include an IPA and plan to grow the range.** IPAs are the most reviewed
   style (Task 4) and their share of all reviews grew from under 5% in 2002
   to 13% in 2011 (chart above); breweries with broader portfolios are rated higher
   (Task 7, correlation 0.35), though partly because good breweries survive
   to grow.
3. **Limitation: this is enthusiast attention, not sales.** RateBeer users
   under-represent the pale-lager drinker who accounts for most beer volume,
   and the data has no prices, costs, or volumes. Use it to pick what wins
   with the enthusiast segment, and validate volume and margin elsewhere.

<!-- Common student variants and what they should find:
     - Boxplot of rating by style instead of bars+SE: fine, same ordering.
     - Rating by ABV with all ABV values: the tail past 15% is noisy (few
       beers); reward students who notice n and cap the axis.
     - Time trend of a single style's average rating: mostly flat within
       style, which is itself a nice contrast with the overall drift (mix
       effect). -->
